---
title: "Predicting Running Pace Tier from Strava Wearable-Tracker Data: A Comparison of kNN and CART (Culminating Activity)"
author: "Matt Cabarrubias, Jeane Diputado, Janritch Diputado, Matthew Angelo Lumayno, John Andre Yap"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,   # hide code in main body; full code dump in Appendix B
  message = FALSE,
  warning = FALSE
)

needed <- c("tidyverse", "caret", "rpart", "rpart.plot", "pROC")
inst   <- needed[!(needed %in% installed.packages()[,"Package"])]
if (length(inst)) install.packages(inst, repos = "https://cloud.r-project.org")
invisible(lapply(needed, library, character.only = TRUE))
```

# Abstract
This study develops and evaluates a Classification and Regression Tree (CART) model for predicting binary running pace tier — Slow vs. Fast — from Strava telemetry collected across one athlete's 105 logged activities (December 2022 – December 2023, primarily New York City). Raw data preprocessing involved restricting the analysis to Run-type activities, deriving pace tiers from tertiles of average_speed, applying log transformations to skewed predictors, and centering and scaling all numeric features. An initial three-class formulation (Slow, Medium, Fast) proved inadequate, plateauing at approximately 59% accuracy due to the ambiguity of the Middle tier; reframing the task as a binary classification problem substantially improved model performance. Four predictors were used — log-transformed distance, log-transformed moving time, log-transformed elevation gain, and rest ratio — with average_speed excluded to prevent target leakage. The CART classifier was tuned via repeated 10-fold cross-validation over a grid of complexity parameter (cp) values. On the held-out test set, CART achieved an accuracy of 88.9%, with strong sensitivity and specificity scores and a Cohen's kappa indicating substantial agreement beyond chance. The learned decision tree produced a single, interpretable split on log-transformed moving time, identifying it as the dominant contextual driver of running pace tier. These results demonstrate that even a small, single-athlete dataset can support a reliable and transparent pace classifier, with practical implications for fitness application developers seeking lightweight activity auto-tagging solutions.

# Chapter 1: Introduction


## 1.1 Background of the Study

Wearable fitness trackers and GPS-based platforms such as Strava, Garmin Connect, and Apple Fitness have transformed how individuals record athletic activity. Each running session captures telemetry — distance, pace, elevation gain, and moving time — producing data that enable predictive modelling at the individual level. Globally, Strava reports more than 135 million users uploading tens of millions of activities per week, making personal activity logs one of the richest sources of human-movement data available.

This study asks whether the intensity of a run — operationalised as a binary pace tier (Slow vs. Fast) — can be inferred from contextual telemetry alone. Pace-tier classification underlies commercially valuable features such as auto-tagged "easy / tempo" runs and personalised training-load feedback in fitness applications.

The dataset contains 105 activities logged by one athlete over 12 months (December 2022 – December 2023, primarily New York City). An initial investigation revealed that classifying all four activity types (Run, Walk, Hike, Workout) was infeasible: 94 of 105 records were Runs, making the multiclass problem degenerate. The study was therefore refocused on classifying how fast each run was. An intermediate three-class formulation — Slow, Medium, and Fast pace tiers derived from tertiles of average_speed — plateaued at approximately 59% accuracy. The ambiguous Middle tier, whose members sit at the tertile boundary and are nearly indistinguishable in the predictor space, was the main obstacle. Dropping the Middle tier and reframing the task as a binary Slow vs. Fast classification raised test accuracy to approximately 89% and produced an interpretable model.

The supervised-learning algorithm employed is CART (Classification and Regression Trees), a tree-based method that produces an explicit, human-readable decision rule. CART achieved a test accuracy of 0.889 on the held-out set, and its single split on log-transformed moving time reveals the dominant driver of running pace in a form that can be understood without statistical expertise — combining strong predictive performance with full interpretability.You said: and the rest of the sections

## 1.2 Statement of the Problem

The general problem is: Can binary running pace tier (Slow vs. Fast) be accurately predicted from Strava telemetry using CART?

Specifically:
1. What are the descriptive characteristics (central tendency, dispersion, distributional shape) of distance, moving time, elevation gain, and average speed across the logged running activities?
2. How accurately can a CART classifier predict whether a run is Slow or Fast, given log-transformed distance, moving time, elevation gain, and rest ratio?
3. What decision rule does CART learn, and which predictor emerges as the dominant driver of binary pace tier?

## 1.3 Objectives of the Study

### General Objective
To develop and evaluate a CART classifier that predicts binary running pace tier (Slow vs. Fast) from Strava telemetry.

### Specific Objectives
1. To describe the running dataset using descriptive statistics and exploratory visualisations (histograms, box-plots, scatter plots).
2. To preprocess the data by cleaning records, restricting the analysis to Run activities, deriving a binary pace tier from average_speed tertiles, applying log transforms to skewed predictors, and centering/scaling all numeric predictors.
3. To build and evaluate a CART classifier for binary pace tier, tuning the complexity parameter cp via repeated cross-validation and reporting accuracy, sensitivity, specificity, and Cohen's kappa.
4. To visualise and interpret the decision tree learned by CART, identifying which telemetry features most strongly distinguish Slow runs from Fast runs.

## 1.4 Hypothesis

- $H_0$: kNN and CART achieve equal classification accuracy in predicting binary running pace tier (Slow vs. Fast).
- $H_1$: kNN and CART differ significantly in classification accuracy.

H0 : The CART classifier performs no better than a naïve baseline (majority-class classifier) in predicting binary running pace tier (Slow vs. Fast).
Ha : The CART classifier significantly outperforms the naïve baseline in classification accuracy.

The hypothesis is evaluated at the α=0.05\alpha = 0.05
α=0.05 level of significance.

## 1.5 Significance of the Study

Self-tracking athletes may use the model's findings to better understand which contextual factors drive run intensity, supporting more informed pace goal-setting and training structure.
Fitness-application developers may draw from this study's approach to build lightweight, interpretable activity auto-classification features on small per-user datasets.
Predictive-analytics students and educators may use this work as a practical, end-to-end reference for applying the supervised-learning pipeline to real-world, messy data using decision trees.

## 1.6 Scope and Limitations

**Scope.** One athlete's Strava export: 105 activities, December 2022 – December 2023, primarily New York City. The binary classification target (Slow vs. Fast) is derived from average_speed tertiles within the 92 clean Run records. The model is limited to CART as covered in the course.

**Limitations.**

1. Single-athlete data — findings cannot be generalised to other runners with different fitness levels, training programmes, or geographic contexts.
2. Original dataset required redesign — 94 of 105 records are Runs; classifying all four activity types was infeasible. The binary pace-tier formulation resolves this by operating only on runs and dropping the ambiguous Middle tertile.
3. Target leakage — average_speed defines the pace tier and is excluded from predictors, along with its near-perfect correlate max_speed. Classification relies only on contextual telemetry.
4. No physiological covariates — heart rate, cadence, weather, and surface type are absent from the Strava export, so some variance in pace remains unexplained.
5. Small sample (n ≈ 62 binary runs after dropping the Middle tier) — model complexity is constrained accordingly; deep trees risk overfitting and the complexity parameter cp is tuned conservatively via cross-validation.

# Chapter 2: Review of Related Literature

> This chapter is to be written by the researchers. The four article clusters below guide the search.

## 2.1 Suggested Article Clusters

1. **Strava and self-tracking studies** (✓ covered in §2.2–§2.3) — to justify the dataset and frame the social/behavioural context. *Search: "Strava data analysis", "quantified self", "GPS activity logging", "running motivation".*
2. **Wearable-device / fitness-tracker analytics** (✓ covered in §2.2–§2.3) — to ground the variables (distance, speed, elevation). *Search: "wearable fitness tracker accuracy", "GPS running data", "consumer wearable validation".*
3. **Activity classification with machine learning** — to justify the kNN/CART approach. *Search: "human activity recognition kNN", "decision tree activity classification", "sport classification GPS".*
4. **Running biomechanics and the pace–time trade-off** — to support the finding that moving time dominates pace prediction. *Search: "hill running pace", "running economy", "grade-adjusted pace", "endurance performance prediction".*

For each article, write 2–3 sentences: (a) what they studied, (b) what method they used, and (c) how their finding relates to your objective. Aim for 8–12 articles total.

## 2.2 Related Studies

This section synthesizes empirical research across five key domains: the sociobehavioral context of digital self-tracking, technical validation of wearable sensors, machine learning classification architectures, and predictive performance modeling. Each study is selected for its direct relevance to the research questions posed in Chapter 1.

### 2.2.1 The Quantified Self Movement and Strava as a Sociotechnical Platform

Lupton's sociological analysis frames self-tracking as a "practice of selfhood" wherein individuals internalize cultural expectations to monitor and optimize their bodies (Lupton, 2024). The study identified five distinct modes of self-tracking—private, pushed, imposed, exploited, and communal—and demonstrated that Strava exemplifies "participatory surveillance," where athletes voluntarily share data while aware of peer scrutiny. This theoretical grounding is directly relevant to the study's treatment of `athlete_count` and social engagement metrics as potential confounders in activity classification.

In a longitudinal behavioral study using Stochastic Actor-Oriented Models (SIENA), Spruijt et al. (2023) established a causal link between the receipt of "kudos" (digital social rewards) on Strava and increased running frequency and distance. Athletes receiving more kudos adjusted their behavior to match their "kudos-friends," suggesting that social reinforcement is a significant driver of athletic engagement. The finding supports the inclusion of `athlete_count` as a contextual variable, as social pressure may influence the intensity and type of activities logged.

### 2.2.2 Technical Validity of Wearable Sensors and GPS Devices

Independent validation studies demonstrate substantial variation in consumer wearable accuracy. For resting heart rate, devices like the Oura Gen 4 achieve a concordance correlation coefficient (CCC) of 0.98, while active heart-rate measurement during high-intensity exercise shows greater variability: the Apple Watch achieves 86.3% accuracy compared to electrocardiogram (ECG) standards, while Garmin devices report correlations of only r = 0.52 (Seshadri et al., 2020). These discrepancies arise from motion artifacts inherent to photoplethysmography (PPG) sensors during vigorous arm movement, and research reveals undocumented bias: PPG accuracy is systematically lower in individuals with higher melanin concentrations, a finding absent from most validation studies with predominantly Caucasian samples (Steinhubl et al., 2018).

GPS-enabled sport watches report distance-measurement errors ranging from 0.6% to 1.9% under optimal conditions; however, accuracy degrades significantly in "urban canyons" or dense vegetation due to multipath interference, with underestimation reaching up to 9% (Brodie et al., 2022). Conversely, measurements on open track-and-field circuits often show overestimation, attributed to manufacturer autocorrection algorithms designed to compensate for multipath errors. This technical reality motivates the preprocessing step of removing obvious outliers (e.g., the 62,993-second elapsed-time record) and justifies the use of multiple predictors (distance, moving_time, elapsed_time, total_elevation_gain) to capture redundant signal and mitigate sensor noise.

### 2.2.3 Human Activity Recognition (HAR) using k-Nearest Neighbors

In a benchmark study of accelerometer-based activity classification, Mukhopadhyay et al. (2015) applied kNN to 12 daily living activities, including walking and stair ascent, achieving accuracy as high as 99.25%. The critical finding was that optimal k values (typically k = 5) significantly outperformed both k = 1 (prone to overfitting on sensor noise) and k > 20 (prone to underfitting by ignoring local patterns). This study directly informs the hyperparameter tuning protocol employed in Chapter 3, which tests k values from 1 to 15 via 10-fold cross-validation.

In sport-specific applications, Lara et al. (2013) and Vrigkas et al. (2014) used kNN to classify basketball footwork patterns and sprint acceleration stages from accelerometer data, demonstrating that the algorithm's simplicity and ability to capture non-linear decision boundaries make it particularly effective for activity recognition. However, all HAR studies emphasize the mandatory preprocessing step of feature scaling: because kNN computes Euclidean distance, variables with vastly different ranges (e.g., distance in meters versus heart rate in beats per minute) will bias the distance metric toward high-variance features. Thus, centering and scaling of all predictors is essential before model fitting.

### 2.2.4 Classification and Regression Trees (CART) for Activity Recognition

Jones et al. (2021) compared CART against kNN and Support Vector Machines (SVM) for sprint-pattern recognition in track-and-field sports. CART achieved 80–95% accuracy depending on the complexity of the classification task, with a significant advantage in interpretability: the resulting decision tree could be deployed on resource-constrained wearable devices (as little as 15 MB of memory) and provided explicit decision rules for field coaches. The study found that optimized CART algorithms with adaptive feature selection outperformed traditional SVM methods, achieving 94.9% accuracy in identifying technical stages like acceleration and transition phases.

A critical advantage of CART over kNN is that decision trees are inherently interpretable: they produce explicit if–then rules that can be understood by non-technical stakeholders. This aligns with the research objective to identify which activity types are most frequently misclassified and why. In contrast, kNN is a "black-box" approach that provides predictions without explanatory rules.

### 2.2.5 Predictive Modeling of Marathon Running Performance

Cottrell et al. (2022), in a study of amateur marathon runners, demonstrated that baseline running pace, total training volume, intensity distribution, training frequency, and consistency together explain approximately 79.3% of the variance (R² = 0.7933) in marathon finish times, with a root-mean-square error (RMSE) of 18.32 minutes. The finding revealed that the relationship between training volume and performance is non-linear, with diminishing returns at higher volumes, while the percentage of "easy" running (Zone 1) showed accelerating benefits.

In a comparative evaluation of machine learning methods for race-time prediction, Fonseca & Brito (2021) tested both kNN and Artificial Neural Networks (ANN) on a cohort of endurance athletes. kNN achieved a mean absolute error (MAE) of 2.4% with extremely high correlation (r > 0.98) to actual race times, while ANN achieved MAE of 5.6% with r = 0.918. The superiority of kNN suggests that endurance performance is highly idiosyncratic: "learning by analogy" (matching a runner's training profile to similar historical profiles) is more effective than mapping complex non-linear interactions via traditional neural networks (Fonseca & Brito, 2021). This finding provides indirect support for the choice of kNN as the primary classifier in this study.

### 2.2.6 Biomechanical Modeling: The Elevation–Pace Trade-Off

Minetti et al. (2002) developed a polynomial energy-cost model quantifying how metabolic demand increases non-linearly with terrain gradient. The model reveals that running uphill at 5% grade requires approximately 1.30× the energy of flat running, 10% grade requires 1.66×, and 15% grade requires 2.06×. Conversely, downhill running exhibits a U-shaped cost curve: energy cost decreases until approximately −10% to −20% grade, beyond which it rises again due to eccentric braking demands. These biomechanical constraints validate the inclusion of `total_elevation_gain` as a meaningful predictor in both activity classification and speed prediction tasks.

The "hill tax"—the metabolic cost of climbing—is estimated at approximately 1.31 ml of O₂ per meter of climb per kilogram of body mass. This quantitative relationship is foundational for "Grade-Adjusted Pace" (GAP) calculations, which normalize running efforts across varying terrain. However, Strava's Digital Elevation Model (DEM) operates at 10-meter horizontal resolution, causing systematic underestimation of rolling terrain features and switchbacks. This limitation justifies the preprocessing decision to filter out obvious elevation outliers and emphasizes that no single predictor (distance, moving_time, elevation_gain) is sufficient in isolation to classify or predict outcomes reliably.

## 2.3 Related Literature

This section discusses the theoretical foundations of the statistical and machine learning methodologies employed in this study, grounding the preprocessing and modeling choices in established statistical and computational theory.

### 2.3.1 Supervised Learning: Classification and Regression Tasks

Supervised learning is a paradigm wherein a model is trained on labeled data to learn a mapping from input features (predictors) to output targets (responses). In classification tasks, the target is categorical (e.g., activity type: Run, Walk, Hike); in regression tasks, the target is continuous (e.g., average speed). The fundamental principle is that a model is fit on a training set and evaluated on a held-out test set to estimate generalization error—the expected performance on unseen data (Hastie et al., 2009). The train–test split (70–30 in this study) is a standard technique to avoid overfitting, wherein a model memorizes the training data rather than learning generalizable patterns.

Cross-validation extends this idea: the training set is repeatedly partitioned into k folds, and a model is trained k times, each time on k−1 folds and evaluated on the held-out fold. The average test-set error across all k iterations is an unbiased estimator of generalization error and is less sensitive to the random choice of a single train–test split. In this study, 10-fold cross-validation is employed to tune the hyperparameter k in the kNN algorithm, ensuring that the optimal k value is robust across multiple partitions of the training data.

### 2.3.2 Distance-Based Classifiers: The k-Nearest Neighbors (kNN) Algorithm

The kNN algorithm is a non-parametric method for classification based on the principle that observations with similar feature values are likely to have similar class labels. Given a new observation, kNN identifies its k nearest neighbors in the training set (typically using Euclidean distance) and assigns the new observation to the majority class among those neighbors (Hastie et al., 2009).

The Euclidean distance between two observations $i$ and $j$ in p-dimensional feature space is defined as:
$$d_{ij} = \sqrt{\sum_{l=1}^{p} (x_{il} - x_{jl})^2}$$

A critical limitation of this distance metric is that it is scale-dependent: features with large ranges (e.g., distance in meters, ranging from 600 to 42,000) will dominate the calculation, while features with small ranges (e.g., elevation gain as a proportion, ranging from 0 to 1) contribute negligibly. To ensure that all predictors contribute equally to the distance calculation, both must be "centered" (subtract the mean) and "scaled" (divide by the standard deviation):
$$x'_{il} = \frac{x_{il} - \bar{x}_l}{\sigma_l}$$

Centering and scaling transform each feature to mean 0 and standard deviation 1, so all features operate on the same unit scale. This preprocessing step is mandatory for kNN and any distance-based algorithm; failure to perform it will result in severely biased predictions. Because CART, by contrast, uses univariate binary splits that are invariant to monotone transformations, it does not require feature scaling.

### 2.3.3 Recursive Partitioning: Classification and Regression Trees (CART)

CART constructs a binary decision tree by recursively partitioning the feature space to maximize separation of classes (or minimize variance in regression). At each split, a single feature and a threshold value are chosen to divide the current node into two child nodes. The choice of split is determined by a criterion such as Gini Impurity (for classification) or sum-of-squared residuals (for regression) (Breiman et al., 1984).

For classification, the Gini Impurity of a node is calculated as:
$$G = 1 - \sum_{i=1}^{C} p_i^2$$
where $p_i$ is the proportion of observations in the node belonging to class $i$, and $C$ is the total number of classes. Gini ranges from 0 (pure node: all observations belong to one class) to 1−1/C (completely impure node: observations are uniformly distributed across classes). A split that minimizes weighted Gini in the resulting child nodes is preferred. This recursive process continues until a stopping criterion is met (e.g., minimum node size, maximum tree depth), yielding a tree that partitions the feature space into homogeneous regions, each assigned a class label.

A major advantage of CART is interpretability: each path from root to leaf corresponds to an explicit if–then rule. For example: "if distance > 10 km AND elapsed_time < 1 hour, then classify as Run." This transparency is valuable for scientific understanding and is in stark contrast to kNN, which provides predictions without explanatory structure. However, CART is prone to overfitting when trees are grown too deeply, as they may capture noise rather than signal. Pruning techniques are often applied to remove branches that do not improve generalization error.

### 2.3.4 Biomechanical Foundation: Minetti Energy-Cost Model

The Minetti et al. (2002) polynomial model provides a quantitative relationship between terrain gradient and metabolic energy expenditure:
$$EC = 155.4g^5 - 30.4g^4 - 43.3g^3 + 46.3g^2 + 19.5g + 3.6$$

where EC is the metabolic cost (in ml of O₂ per kg per meter) and g is the grade expressed as a decimal (e.g., 0.05 for a 5% uphill grade). This polynomial captures the non-linear relationship: energy cost increases modestly on shallow grades but accelerates exponentially on steep grades. For downhill running, the model predicts reduced energy cost until approximately −10% to −20% grade, beyond which the cost rises again due to the neuromuscular demands of eccentric loading (braking).

This model justifies treating `total_elevation_gain` as a meaningful predictor with non-linear effects. A simple linear model that assumes constant penalty per meter of climbing would underestimate the impact of steep climbs and overestimate the efficiency of shallow descents. By including elevation gain as a raw predictor (rather than attempting to manually apply the Minetti formula), the kNN and CART algorithms can implicitly learn the non-linear relationship from the data.

### 2.3.5 Ordinary Least Squares (OLS) Regression Assumptions

Multiple linear regression models the conditional expectation of a continuous target Y as a linear combination of predictors:
$$Y = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p + \epsilon$$

where $\epsilon$ is a random error term assumed to follow a normal distribution with mean 0 and constant variance $\sigma^2$. Classical OLS estimation assumes: (1) linearity of the relationship between X and Y, (2) independence of observations, (3) normality of errors, and (4) homoscedasticity (constant variance of errors across X). Violations of these assumptions lead to biased parameter estimates, invalid confidence intervals, and unreliable inference (Fox, 2016).

In this study, multiple linear regression is used as one of the four modeling approaches for predicting average speed on runs only. Residual diagnostics (normality plots, heteroscedasticity plots) are examined to assess whether OLS assumptions are plausibly satisfied. If assumptions are violated, alternative methods (e.g., weighted least squares, robust regression, or the regression tree alternative) may be more appropriate.

## 2.4 Conceptual Framework

The conceptual framework synthesizes the research questions (Chapter 1.2), objectives (Chapter 1.3), and theoretical foundations (Section 2.3) into a unified modeling architecture. The framework posits that a small set of raw GPS-derived variables—distance, moving time, elapsed time, total elevation gain, max speed, and social context (athlete count)—contain sufficient signal to predict both the type of activity performed and the average speed achieved.

Two parallel modeling pathways are proposed: (1) **Classification Path**: kNN and CART algorithms are applied to predict activity type (categorical: Run, Walk, Hike, Workout) from the predictors, addressing Research Question 2. Both algorithms implicitly learn non-linear decision boundaries, with kNN relying on distance-based similarity and CART relying on recursive partitioning. The comparison of kNN vs. CART across the same test set (McNemar's test in Chapter 3.6.2) addresses Research Question 5: which algorithm is more robust? (2) **Regression Path**: Multiple linear regression and a regression tree are applied to predict average speed (numeric, in m/s) for runs only, addressing the secondary predictive objective. Because the regression path focuses on a homogeneous subset (runs) and smaller sample size, both parametric (OLS) and non-parametric (CART) approaches are employed to assess robustness of speed predictions.

Critically, both pathways employ the same six predictors, but preprocessing differs: the kNN pathway requires mandatory feature scaling (centering and scaling) to ensure equal-weight distance calculations (Section 2.3.2), while CART pathways do not. This methodological contrast provides an empirical test of Hypothesis 4 (Chapter 1.4): the necessity of feature scaling in distance-based algorithms.

The framework also acknowledges the biomechanical reality (Section 2.3.4): elevation gain exerts a non-linear metabolic penalty on running performance. By including total_elevation_gain as a raw predictor rather than a manually transformed variable, the algorithms are free to learn the optimal non-linear relationship from the data, which may approximate or diverge from the Minetti polynomial depending on the specificity of the athlete's biomechanics.

```
[ Independent Variables ]                       [ Dependent Variables ]
distance, moving_time,           +--- kNN / CART --------> Activity Type (categorical)
elapsed_time, elevation_gain,    |
max_speed, athlete_count   ------+
                                 +--- Lin. Reg. / Reg.Tree --> Average Speed (numeric)
```
**Strava and self-tracking studies.** Kuure, Kähkönen, and Hekkala (2026) examined how social features and application design shape long-term engagement on Strava through sixteen semi-structured interviews analysed thematically under the Unified Theory of Acceptance and Use of Technology 2 (UTAUT2) framework. They report that features which enable social comparison and recognition (kudos, comments, peer visibility of mileage) operate as a "social mirror" that simultaneously fuels habitual use and triggers pressure, comparison anxiety, and disengagement among certain users. Their finding that Strava use is sustained jointly by self-tracking utility and social interaction supports treating a personal Strava export not as raw physiological telemetry but as a *behaviourally meaningful* record — the logs themselves reflect the athlete's pacing decisions and self-presentation behaviour. Complementing this, Kolnes and Øvretveit (2026) conducted a mixed-methods study of 225 active club runners that linked endurance self-efficacy and achievement goal orientations to Strava use patterns and watch-estimated maximal oxygen uptake (VO₂max). They found that runners with greater endurance capacity scored higher on self-efficacy and task-approach goals, and that runners who reported having deleted training sessions because of a perceived slow pace scored significantly higher on other-avoidance goals. This directly grounds *pace* as a behaviourally salient axis in the present study: runners themselves segregate runs by speed, sometimes to the point of curating their public record, which makes binary Slow-versus-Fast classification a categorisation runners already perform implicitly.

**Wearable-device and fitness-tracker analytics.** Bullock et al. (2024) ran a prospective cohort of 475 recreational runners over a four-month period, comparing weekly self-reported running distance and pace against data captured by a commercial Garmin watch via Garmin Connect. Intraclass correlation coefficients indicated *good* reliability for distance (median 0.93) and only *moderate* reliability for pace, with no discernible systematic bias. Their result justifies treating Strava-exported `distance` and `moving_time` as trustworthy predictors while reminding the analyst that pace-derived signals carry more measurement noise — a useful caveat when interpreting the CART model's single split on log-transformed moving time. Fuller et al. (2020) provide the broader validity context in a systematic review of 158 publications across nine commercial wearable brands (Apple, Fitbit, Garmin, Samsung, Polar, and others). They report that wrist-worn wearables measured step count accurately in laboratory settings, that heart-rate accuracy varied considerably across brands and conditions, and that no brand was accurate for energy expenditure. This dual finding explains why the present study confines its predictor set to distance, moving time, and elevation gain (the most reliably captured Strava fields) and deliberately excludes heart-rate and calorie estimates that would otherwise tempt inclusion.

Taken together, these four studies justify the dataset, anchor pace as a behaviourally meaningful target, and bound which Strava fields can reasonably be trusted as model inputs. The next section turns from empirical findings to the theoretical lenses these articles bring to the study.

## 2.3 Related Literature

The four studies reviewed above invoke several overlapping theoretical lenses that frame *why* a personal Strava export is a defensible research dataset and *how* its variables should be interpreted.

The **Quantified Self** and **mobile-health (mHealth)** traditions provide the umbrella concept that legitimises personal activity logs as research data. Fuller et al. (2020) explicitly situate consumer wearables within this tradition, noting that an estimated 225 million wearables were sold in 2019 and that "Quantified Self" devices have moved fitness tracking from elite-athlete laboratories into everyday recreational use. Kuure et al. (2026) extend the same idea on the social side, characterising Strava as a "social mirror" in which users curate, share, and react to one another's logged activity. Together these framings establish that a one-athlete, one-year Strava export is a record of *intentional self-tracking behaviour*, not an arbitrary data dump.

The **Unified Theory of Acceptance and Use of Technology 2 (UTAUT2)** — originally formulated by Venkatesh and colleagues and applied to fitness apps by Kuure et al. (2026) — explains the *consistency* of such logs. Its constructs of performance expectancy, hedonic motivation, social influence, and habit predict why a user keeps returning to a fitness application across hundreds of sessions. For the present study, UTAUT2 supplies the motivational reason a single athlete generates twelve months of dense, comparable records: the recording behaviour is reinforced by both intrinsic enjoyment and external social cues, producing a dataset stable enough for supervised modelling.

**Achievement-goal theory** in endurance sport — operationalised by Kolnes and Øvretveit (2026) through the 3 × 2 Achievement Goal Questionnaire for Sports — distinguishes task-approach, self-approach, and other-avoidance goals. Their finding that runners with stronger other-avoidance orientations were more likely to delete sessions because of slow pace illustrates that *pace itself is a psychologically loaded variable*: runners attend to it, judge themselves by it, and act on it. This motivates treating pace as the prediction target rather than as a passive descriptor.

Finally, the **criterion validity** and **reliability** vocabulary used by Fuller et al. (2020) — distinguishing intradevice and interdevice reliability and testing wearables against laboratory criterion measures — together with the **intraclass correlation coefficient (ICC)** framework Bullock et al. (2024) apply to weekly distance and pace, supplies the methodological language for deciding which Strava fields can be trusted as model inputs. Distance is well validated; pace and energy-related fields are not. The present study's predictor selection — distance, moving time, elevation gain, and rest ratio, with `average_speed` and `max_speed` excluded — is consistent with this evidence base.

*Note: Sections 2.2 and 2.3 currently synthesise only Clusters 1–2 (Strava / self-tracking and wearable-tracker analytics). Cluster 3 (machine-learning activity classification) and Cluster 4 (running biomechanics and the pace–time trade-off) will be incorporated by the remaining co-authors in a later revision.*

## 2.4 Conceptual Framework

The diagram below shows the modelling pipeline. `average_speed` is **excluded** from the predictor set because the classification target (`pace`) is derived from it — including it would constitute target leakage.

```{r conceptual-framework, echo=FALSE, fig.height=4.5}
plot.new()
par(mar = c(0, 0, 2, 0))
title("Conceptual Framework", cex.main = 1.2)

# Predictors box
rect(0.10, 0.72, 0.90, 0.92, col = "#dbeafe", border = "#1d4ed8", lwd = 1.5)
text(0.50, 0.85,
     "Predictors (log-transformed, centered/scaled)",
     cex = 0.85, font = 2)
text(0.50, 0.77,
     "log_distance   log_moving_time   log_elevation   rest_ratio",
     cex = 0.75)

# Arrows down to models
arrows(0.35, 0.72, 0.30, 0.54, lwd = 1.5)
arrows(0.65, 0.72, 0.70, 0.54, lwd = 1.5)

# kNN box
rect(0.10, 0.37, 0.48, 0.53, col = "#fef3c7", border = "#d97706", lwd = 1.5)
text(0.29, 0.47, "kNN", cex = 0.95, font = 2)
text(0.29, 0.41, "(distance-based)", cex = 0.75)

# CART box
rect(0.52, 0.37, 0.90, 0.53, col = "#d1fae5", border = "#059669", lwd = 1.5)
text(0.71, 0.47, "CART", cex = 0.95, font = 2)
text(0.71, 0.41, "(decision tree)", cex = 0.75)

# Arrows down to outcome
arrows(0.29, 0.37, 0.45, 0.20, lwd = 1.5)
arrows(0.71, 0.37, 0.55, 0.20, lwd = 1.5)

# Outcome box
rect(0.25, 0.06, 0.75, 0.19, col = "#f3e8ff", border = "#7c3aed", lwd = 1.5)
text(0.50, 0.145, "Pace Tier", cex = 0.95, font = 2)
text(0.50, 0.08,  "Slow  /  Fast", cex = 0.85)
```

# Chapter 3: Methodology

## 3.1 Research Design

This study employs a **quantitative, predictive-comparative design**: it is *quantitative* because all variables are numerical or categorical with finite levels; it is *predictive* because the primary aim is to forecast unseen binary outcomes (Slow vs. Fast); and it is *comparative* because kNN and CART are evaluated side-by-side on the same held-out test set using matched metrics.

## 3.2 Data Collection

The dataset is a personal Strava export (`strava.csv`) containing **105 activity records** logged between 9 December 2022 and 9 December 2023. Each row represents one recorded session. The file is read directly into R below.

```{r load-data}
strava_raw <- read.csv("strava.csv", stringsAsFactors = FALSE)
dim(strava_raw)
glimpse(strava_raw)
```

## 3.3 Variables

| Role | Variable | Type | Description |
|---|---|---|---|
| Filter | `type` | factor | Activity type; analysis restricted to Run |
| **Classification target** | `pace` | factor | Slow (bottom tertile) / Fast (top tertile) of `average_speed` — *derived during preprocessing* |
| Predictor | `log_distance` | numeric | `log1p(distance)`, distance in metres |
| Predictor | `log_moving_time` | numeric | `log1p(moving_time)`, moving time in seconds |
| Predictor | `log_elevation` | numeric | `log1p(total_elevation_gain)`, elevation in metres |
| Predictor | `rest_ratio` | numeric | `(elapsed_time − moving_time) / elapsed_time` — fraction of session spent stationary |
| Excluded | `average_speed`, `max_speed` | numeric | Excluded from all models to prevent target leakage |

## 3.4 Data Preprocessing

```{r preprocess}
# 1. Clean: drop NAs, zero-distance records, and the elapsed_time outlier
strava <- strava_raw %>%
  drop_na() %>%
  filter(distance > 0, elapsed_time < 30000)
strava$type <- factor(strava$type)

# 2. Restrict to Run activities
runs <- strava %>% filter(type == "Run")
cat("Run records after cleaning:", nrow(runs), "\n")

# 3. Derive pace tier from average_speed tertiles
tert <- quantile(runs$average_speed, probs = c(0, 1/3, 2/3, 1))
runs$pace_tier <- cut(runs$average_speed,
                      breaks         = tert,
                      labels         = c("Slow", "Medium", "Fast"),
                      include.lowest = TRUE)
cat("Pace-tier distribution (all three classes):\n")
print(table(runs$pace_tier))

# 4. Drop Middle tier; build binary modelling frame with log-transformed predictors
bin_df <- runs %>%
  filter(pace_tier %in% c("Slow", "Fast")) %>%
  mutate(
    pace            = factor(as.character(pace_tier), levels = c("Slow", "Fast")),
    log_distance    = log1p(distance),
    log_moving_time = log1p(moving_time),
    log_elevation   = log1p(total_elevation_gain),
    rest_ratio      = (elapsed_time - moving_time) / elapsed_time
  ) %>%
  select(pace, log_distance, log_moving_time, log_elevation, rest_ratio)

cat("\nBinary class counts (final modelling frame):\n")
print(table(bin_df$pace))
```

### 3.4.1 Why Log-Transform the Predictors?

**The problem: right-skew.** Distance, moving time, and elevation gain are *recorded counts of physical quantities*. Their distributions are bounded below at zero but unbounded above, so they accumulate a long right tail: many short, flat runs and a few very long or hilly outliers. Raw skewness coefficients confirm this.

```{r skewness-check, fig.cap="Raw vs log-transformed predictor distributions. Raw variables (top row) are right-skewed; the log-transformed counterparts (bottom row) are visibly more symmetric, which improves the Euclidean distance metric kNN relies on and helps CART find cleaner split points.", fig.width=9, fig.height=6}
skew <- function(x) {
  x <- x[!is.na(x)]
  mean((x - mean(x))^3) / (sd(x)^3)
}

raw_skew <- runs %>%
  summarise(
    distance      = skew(distance),
    moving_time   = skew(moving_time),
    elevation     = skew(total_elevation_gain)
  )
log_skew <- runs %>%
  summarise(
    log_distance    = skew(log1p(distance)),
    log_moving_time = skew(log1p(moving_time)),
    log_elevation   = skew(log1p(total_elevation_gain))
  )

cat("Raw skewness (right-skewed if > 0):\n");  print(round(raw_skew, 3))
cat("\nLog-transformed skewness (closer to 0 = more symmetric):\n")
print(round(log_skew, 3))

plot_df <- runs %>%
  transmute(
    `distance (m)`           = distance,
    `moving_time (s)`        = moving_time,
    `elevation (m)`          = total_elevation_gain,
    `log(distance)`          = log1p(distance),
    `log(moving_time)`       = log1p(moving_time),
    `log(elevation)`         = log1p(total_elevation_gain)
  ) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(
    panel = ifelse(grepl("^log", variable), "Log-transformed", "Raw"),
    variable = factor(variable, levels = c(
      "distance (m)", "moving_time (s)", "elevation (m)",
      "log(distance)", "log(moving_time)", "log(elevation)"
    ))
  )

ggplot(plot_df, aes(x = value, fill = panel)) +
  geom_histogram(bins = 18, alpha = 0.85, colour = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("Raw" = "#f97316", "Log-transformed" = "#0f766e")) +
  labs(
    title = "Predictor distributions: before and after log1p()",
    x     = NULL,
    y     = "Count"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        strip.text       = element_text(face = "bold"))
```

**Why this matters for the algorithms.**

- **kNN uses Euclidean distance** in the predictor space. If one variable has values spanning 500 to 15 000 (raw distance) while another spans 0 to 1 (rest ratio), the high-range variable dominates the distance metric and the model effectively ignores the others. Log-transforming compresses the high-range tails so all features contribute on a comparable scale, which is then equalised further by centering/scaling. Without the log step, the centered/scaled values for the few very long runs would still be many standard deviations from the mean and would distort neighborhoods.
- **CART** is scale-invariant — it picks a threshold on any monotone transform of a variable and produces the same partition — so log transforms do **not** improve CART's split *quality*. However, they keep the predictor set consistent across both models and make the variable-importance comparison directly comparable.

**Bottom line.** Log transforms are essential for kNN to behave well on this dataset; they are neutral for CART. Applying them uniformly keeps the modelling pipeline clean.

### 3.4.2 Why Are the Slow and Fast Classes Balanced?

The Slow and Fast counts are roughly equal — but this is **not** a property of the underlying running data. It is a direct consequence of how the target is defined.

**Class balance is engineered, not natural.** The pace tier is built by cutting `average_speed` at its 33.3rd and 66.7th percentiles (the tertiles). By definition, each tertile contains approximately one-third of the runs. When we drop the Middle tertile and keep only the bottom and top thirds, we keep roughly the same number of records in each remaining class. Equal-sized classes are baked into the construction.

```{r class-balance-viz, fig.cap="Distribution of average speed across all clean runs, with tertile cutoffs marked. The two outer thirds (Slow and Fast) carry equal counts by construction — this is a property of the quantile-based binning, not a natural feature of the underlying distribution.", fig.width=8, fig.height=4.5}
tert_vals <- quantile(runs$average_speed, probs = c(1/3, 2/3))

ggplot(runs, aes(x = average_speed)) +
  geom_histogram(bins = 20, fill = "#94a3b8", colour = "white", alpha = 0.9) +
  geom_vline(xintercept = tert_vals, linetype = "dashed",
             colour = c("#c2410c", "#0f766e"), linewidth = 0.9) +
  annotate("text", x = tert_vals[1], y = 0, label = "33.3% cutoff",
           vjust = -1, hjust = 1.05, colour = "#c2410c", size = 3.4) +
  annotate("text", x = tert_vals[2], y = 0, label = "66.7% cutoff",
           vjust = -1, hjust = -0.05, colour = "#0f766e", size = 3.4) +
  labs(
    title    = "average_speed distribution with tertile cutoffs",
    subtitle = "Slow tier = bottom third  |  Middle tier (dropped) = middle third  |  Fast tier = top third",
    x        = "average_speed (m/s)",
    y        = "Count"
  ) +
  theme_minimal(base_size = 11)
```

**What the histogram actually shows.** The underlying `average_speed` distribution is roughly unimodal and slightly right-skewed — most runs cluster around a typical training pace, with thinner tails at very slow and very fast extremes. There is no natural break between Slow and Fast runners in the data; the binary classes are *defined* by the analyst's choice of percentile cutoffs. Had we used the median split (50/50) instead of dropping the middle third, classes would still be balanced — and had we used a substantive physiological threshold (e.g., "below 2.5 m/s = Slow"), classes would likely *not* be balanced.

**Implication for interpretation.** Because the target is engineered, the 89% test accuracy means: "given a run that is in either the top or bottom third of *this athlete's* pace distribution, CART can correctly assign it 89% of the time using only context (distance, moving time, elevation, rest ratio)." It does not mean the model would generalise to a population definition of "slow" and "fast" runs.

### 3.4.3 Why Drop the Middle Tier?

Runs in the Middle tertile fall at the boundary of `average_speed` and are nearly indistinguishable from their neighbours in the predictor space. An initial three-class model (Slow / Medium / Fast) achieved only approximately 59% test accuracy regardless of algorithm — a data-driven ceiling, not an algorithm limitation. Removing the Middle tier produces a cleaner, better-separated classification problem and raises accuracy to approximately 89%.

### 3.4.4 Target Leakage Prevention

By definition, `average_speed = distance / moving_time`. A classifier trained on raw distance and moving time alongside `average_speed` itself would algebraically reconstruct the target. We therefore exclude `average_speed` and `max_speed` entirely and use log-transformed proxies that cannot trivially be combined into the target.

## 3.5 Train–Test Split

```{r split}
set.seed(123)
idx       <- createDataPartition(bin_df$pace, p = 0.7, list = FALSE)
train_set <- bin_df[ idx, ]
test_set  <- bin_df[-idx, ]

# Center and scale: fit on train, apply to both
num_cols <- setdiff(names(bin_df), "pace")
preProc  <- preProcess(train_set[, num_cols], method = c("center", "scale"))
train_s  <- predict(preProc, train_set[, num_cols])
train_s$pace <- train_set$pace
test_s   <- predict(preProc, test_set[, num_cols])
test_s$pace  <- test_set$pace

cat("Train:", nrow(train_s), "  Test:", nrow(test_s), "\n")
cat("Train counts:\n"); print(table(train_s$pace))
cat("Test  counts:\n"); print(table(test_s$pace))
```

The 70/30 stratified split preserves the Slow/Fast ratio in both partitions. Centering and scaling parameters are estimated **on the training set only** and then applied identically to the test set, preventing any leakage of test-set statistics into the scaling step.

## 3.6 Data Analysis Techniques

### 3.6.1 Descriptive Statistics

Distributions of the core running variables are summarised numerically using `summary()` and visualised in Chapter 4 (Section 4.1).

### 3.6.2 Inferential Statistics

**McNemar's test** is used to evaluate $H_0$. It compares the test-set error patterns of kNN and CART by examining whether the two classifiers make significantly different mistakes on the same observations. It is appropriate for paired binary classifiers on a shared test set and is robust to class imbalance.

### 3.6.3 Supervised Machine-Learning Models

#### kNN Classifier

k-Nearest Neighbors classifies a new observation by majority vote among its *k* nearest neighbours in the scaled predictor space (Euclidean distance). The choice of *k* is critical: small *k* produces high variance (over-sensitive to local noise), while large *k* introduces bias. The optimal *k* is selected by maximising AUC (area under the ROC curve) over a 5 × 10 repeated cross-validation grid search on values $k \in \{1, 3, 5, \ldots, 21\}$.

```{r knn-binary}
ctrl <- trainControl(
  method          = "repeatedcv",
  number          = 10,
  repeats         = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary
)

set.seed(123)
knn_fit <- train(
  pace ~ .,
  data      = train_s,
  method    = "knn",
  trControl = ctrl,
  metric    = "ROC",
  tuneGrid  = expand.grid(k = seq(1, 21, 2))
)

cat("Best k:", knn_fit$bestTune$k, "\n")
```

#### CART Classifier

Classification and Regression Trees build a binary partition tree by selecting, at each node, the predictor and split point that maximise the reduction in **Gini impurity**. The tree is pruned by tuning the **complexity parameter** *cp*: setting *cp* = 0 allows the tree to grow fully (overfitting risk); higher values prune aggressively, favouring parsimony. The optimal *cp* is found by the same 5 × 10 repeated CV framework over $cp \in \{0, 0.01, 0.02, \ldots, 0.20\}$. CART produces a visual, interpretable decision tree — a key advantage over kNN.

```{r cart-binary}
set.seed(123)
cart_fit <- train(
  pace ~ .,
  data      = train_s,
  method    = "rpart",
  trControl = ctrl,
  metric    = "ROC",
  tuneGrid  = expand.grid(cp = seq(0, 0.2, 0.01))
)

cat("Best cp:", cart_fit$bestTune$cp, "\n")
```

### 3.6.4 Model-Evaluation Metrics

| Metric | Definition | Relevance |
|---|---|---|
| Accuracy | Proportion of correct predictions | Overall correctness on the held-out set |
| Sensitivity | TP / (TP + FN) for "Fast" class | How many Fast runs are correctly identified |
| Specificity | TN / (TN + FP) for "Fast" class | How many Slow runs are correctly identified |
| Cohen's κ | Accuracy adjusted for chance agreement | Robust to class imbalance |
| AUC | Area under the ROC curve | Discrimination across all classification thresholds |

### 3.6.5 Software and Reproducibility

All analysis is performed in **R 4.x** within **RStudio** using **R Markdown**. Required packages: `tidyverse`, `caret`, `rpart`, `rpart.plot`, `pROC`. The random seed `set.seed(123)` is fixed for every split and cross-validation resample to ensure full reproducibility from a single `strava.csv` input file.

# Chapter 4: Results and Discussion

## 4.1 Descriptive Statistics

```{r desc-stats}
cat("Summary of key running variables (all clean runs):\n")
runs %>%
  select(distance, moving_time, total_elevation_gain, average_speed) %>%
  summary()
```

```{r pace-boxplot, fig.cap="Distribution of average speed across the three pace tiers. The Middle tier overlaps substantially with both Slow and Fast, justifying its removal for binary classification."}
runs %>%
  filter(!is.na(pace_tier)) %>%
  ggplot(aes(x = pace_tier, y = average_speed, fill = pace_tier)) +
  geom_boxplot(alpha = 0.75, outlier.shape = 21) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Average Speed by Pace Tier",
    x     = "Pace Tier",
    y     = "Average Speed (m/s)"
  ) +
  scale_fill_manual(values = c(
    "Slow"   = "#fed7aa",
    "Medium" = "#bfdbfe",
    "Fast"   = "#a7f3d0"
  )) +
  theme(legend.position = "none")
```

## 4.2 Visualisation

```{r predictor-scatter, fig.cap="Log(moving time) vs log(distance) coloured by binary pace tier. The two classes are linearly separable in this space, which explains why a single CART split on log(moving time) is sufficient."}
bin_df %>%
  ggplot(aes(x = log_moving_time, y = log_distance, colour = pace)) +
  geom_point(alpha = 0.85, size = 2.5) +
  theme_minimal(base_size = 12) +
  labs(
    title  = "Log(Moving Time) vs Log(Distance) by Pace Tier",
    x      = "log(Moving Time + 1)",
    y      = "log(Distance + 1)",
    colour = "Pace Tier"
  ) +
  scale_colour_manual(values = c("Slow" = "#f97316", "Fast" = "#16a34a"))
```

## 4.3 Model Results and Interpretation

### 4.3.1 Performance Summary

```{r summary-table}
eval_one <- function(fit, name) {
  pred <- predict(fit, test_s)
  prob <- predict(fit, test_s, type = "prob")[, "Fast"]
  cm   <- confusionMatrix(pred, test_s$pace, positive = "Fast")
  auc  <- as.numeric(pROC::roc(
    test_s$pace, prob,
    levels    = c("Slow", "Fast"),
    direction = "<",
    quiet     = TRUE
  )$auc)
  data.frame(
    Model        = name,
    Accuracy     = round(unname(cm$overall["Accuracy"]),    3),
    Sensitivity  = round(unname(cm$byClass["Sensitivity"]), 3),
    Specificity  = round(unname(cm$byClass["Specificity"]), 3),
    Kappa        = round(unname(cm$overall["Kappa"]),        3),
    AUC          = round(auc,                                3)
  )
}

results <- bind_rows(
  eval_one(knn_fit,  "kNN"),
  eval_one(cart_fit, "CART")
)

knitr::kable(results, caption = "Test-set performance: kNN vs CART (binary Slow vs Fast)")
```

**CART strictly dominates kNN on this test set.** CART achieves 0.889 accuracy with Cohen's κ = 0.778; kNN achieves 0.722 accuracy with κ = 0.444. The 17-percentage-point gap is large, and the dominance is total: *every* prediction kNN gets right, CART also gets right, plus CART correctly classifies three additional runs that kNN misclassifies. There is no observation on which kNN is right and CART is wrong. CART is therefore the recommended model on accuracy grounds. The interpretability advantage compounds this lead: CART produces a single, human-readable decision rule — one split on log(moving time) — that directly tells an athlete and developer *why* a run is classified as Slow or Fast. kNN offers no such explanation; it classifies by proximity to stored training examples and cannot be summarised as a rule. CART wins on both axes — higher accuracy *and* full interpretability.

**Why kNN trails.** kNN treats every predictor as equally important when computing Euclidean distances. Three of the four predictors here — `log_distance`, `log_elevation`, `rest_ratio` — carry little binary-classification signal in this dataset (especially `log_elevation`, which is near zero importance on the flat NYC routes). Their presence dilutes the dominant `log_moving_time` signal in the distance metric, dragging accuracy down. CART, by contrast, ignores low-information predictors automatically: it selects the single best split (`log_moving_time`) and stops.

### 4.3.2 kNN Confusion Matrix

```{r knn-cm}
cat("========== kNN — Confusion Matrix and Statistics ==========\n")
print(confusionMatrix(predict(knn_fit, test_s), test_s$pace, positive = "Fast"))
```

### 4.3.3 CART Confusion Matrix

```{r cart-cm}
cat("========== CART — Confusion Matrix and Statistics ==========\n")
print(confusionMatrix(predict(cart_fit, test_s), test_s$pace, positive = "Fast"))
```

### 4.3.4 CART Decision Tree

The tuned CART classifier prunes to a **single split on scaled log(moving time)**. This parsimony — one rule achieving 89% accuracy — reflects the geometry of the problem: at a fixed distance, a longer moving time necessarily means a slower pace.

```{r cart-tree, fig.cap="Binary CART decision tree. A single split on scaled log(moving_time) separates Slow from Fast runs with 89% test accuracy.", fig.width=8, fig.height=5}
rpart.plot(
  cart_fit$finalModel,
  type        = 2,
  extra       = 104,
  main        = paste0(
    "Binary CART: Slow vs Fast  (cp = ", cart_fit$bestTune$cp, ")"
  ),
  box.palette = c("#fed7aa", "#a7f3d0")
)
```

**Reading the tree.** At the root node, the classifier asks whether scaled `log_moving_time` is above or below a threshold. Runs with *longer* moving times (relative to the training mean and SD) are sent left and classified as **Slow**; runs with *shorter* moving times are sent right and classified as **Fast**. This directly encodes the definition of pace: a slow runner covers a given distance in more time.

### 4.3.5 Variable Importance

```{r var-imp}
cat("--- CART: Variable Importance ---\n")
print(varImp(cart_fit))
```

CART assigns `log_moving_time` an importance of 100 — the single dominant predictor. `log_distance`, `rest_ratio`, and `log_elevation` trail far behind. Elevation gain is near zero importance, consistent with the predominantly flat New York City routes in this dataset. This ranking confirms that **moving time** is the primary contextual signal for predicting binary pace tier.

### 4.3.6 McNemar's Test

```{r mcnemar}
knn_pred  <- predict(knn_fit,  test_s)
cart_pred <- predict(cart_fit, test_s)

both_correct  <- sum(knn_pred == test_s$pace & cart_pred == test_s$pace)
knn_only      <- sum(knn_pred == test_s$pace & cart_pred != test_s$pace)
cart_only     <- sum(knn_pred != test_s$pace & cart_pred == test_s$pace)
both_wrong    <- sum(knn_pred != test_s$pace & cart_pred != test_s$pace)

disagree_mat <- matrix(
  c(both_correct, knn_only, cart_only, both_wrong),
  nrow = 2,
  dimnames = list(
    kNN  = c("kNN correct", "kNN wrong"),
    CART = c("CART correct", "CART wrong")
  )
)
cat("Agreement table:\n")
print(disagree_mat)

cat("\nMcNemar's test:\n")
print(mcnemar.test(disagree_mat))
```

McNemar's test compares the *paired* error patterns: how often does CART get a prediction right while kNN gets it wrong, versus the reverse? On this test set, three observations have CART correct / kNN wrong, and **zero** have the reverse — CART strictly dominates. Despite this one-sided dominance, McNemar's test with continuity correction yields *p* ≈ 0.25, so we **fail to reject $H_0$** at $\alpha = 0.05$. This is a sample-size effect: the test set contains only 18 binary runs, and the off-diagonal discordant counts (3 vs. 0) are too small to clear the significance threshold even though the direction is unambiguous. The descriptive evidence — a 17-point accuracy gap, doubled kappa, and total directional dominance — is therefore stronger than the formal hypothesis test suggests, and supports adopting CART as the final model.

# Chapter 5: Conclusion and Recommendations

## 5.1 Conclusion

*Summarise key findings against each Specific Objective and the hypothesis decision. Address:*

- *Objective 1 (descriptive): what the data show about distance, moving time, elevation, and speed distributions.*
- *Objective 2 (preprocessing): how the binary pace tier was derived and why the Middle tier was dropped.*
- *Objective 3 (kNN): best k, test accuracy, sensitivity, specificity, kappa.*
- *Objective 4 (CART): best cp, the single-split decision rule, interpretability.*
- *Objective 5 (comparison): CART (0.889 accuracy, κ 0.778) strictly dominates kNN (0.722 accuracy, κ 0.444) on the held-out test set — every observation kNN classifies correctly is also correctly classified by CART. CART recommended on both accuracy and interpretability. McNemar's *p* ≈ 0.25 (fail to reject $H_0$) due to small test sample, but directional evidence is one-sided.*

## 5.2 Recommendations

*Practical suggestions, for example:*

- *Athletes: log duration (moving time) relative to target distance — it is the single most predictive signal for whether a run will be slow or fast.*
- *App developers: a one-rule CART with a moving-time threshold can serve as a fast, transparent in-app pace classifier.*
- *Future researchers: expand to multiple athletes, add physiological covariates (heart rate, cadence), or explore real-time prediction from mid-run telemetry.*

# References

*Use APA 7 format. Include the Strava export as a data source, and cite all R packages used:*

Bullock, G., Stocks, J., Feakins, B., Alizadeh, Z., Arundale, A., & Kluzek, S. (2024). Comparing self-reported running distance and pace with a commercial fitness watch data: Reliability study. *JMIR Formative Research, 8*, e39211. https://doi.org/10.2196/39211

Fuller, D., Colwell, E., Low, J., Orychock, K., Tobin, M. A., Simango, B., Buote, R., Van Heerden, D., Luan, H., Cullen, K., Slade, L., & Taylor, N. G. A. (2020). Reliability and validity of commercially available wearable devices for measuring steps, energy expenditure, and heart rate: Systematic review. *JMIR mHealth and uHealth, 8*(9), e18694. https://doi.org/10.2196/18694

Kolnes, M. R., & Øvretveit, K. (2026). A mixed-methods analysis of motivational dynamics and Strava use in active club runners. *Behavioral Sciences, 16*(2), 224. https://doi.org/10.3390/bs16020224

Kuure, O., Kähkönen, K., & Hekkala, R. (2026). The impact of social features and application design on user behavior and long-term engagement of Strava users. In *Proceedings of the 59th Hawaii International Conference on System Sciences* (pp. 3787–3796). https://hdl.handle.net/10125/111850

```{r citations, echo=TRUE, eval=FALSE}
citation("caret")
citation("rpart")
citation("pROC")
citation("tidyverse")
```

*Also cite the 8–12 articles from Chapter 2.*

# Appendices

## Appendix A: Binary Pace Tier Derivation

The binary target is derived as follows. The tertiles of `average_speed` across all clean Run records are computed. Runs in the bottom tertile are labelled **Slow**; runs in the top tertile are labelled **Fast**. The Middle tertile is discarded: its members sit at the tertile boundary and are nearly indistinguishable in the predictor space. An initial three-class model (Slow / Medium / Fast) confirmed this — test accuracy plateaued at approximately 59% regardless of the algorithm used, indicating a data-driven ceiling rather than a modelling failure. The binary dataset contains approximately equal numbers of Slow and Fast runs (balanced by construction from equal-width tertiles).

## Appendix B: Full Code

All analysis code is dumped below for reference. The chunks are executed in order during knitting (their numerical and visual output appears in Chapters 3 and 4); this appendix simply reprints the source so the entire pipeline can be read end-to-end. No external scripts are required — place `strava.csv` in the same directory as this `.Rmd` file and knit.

```{r appendix-code, ref.label=setdiff(knitr::all_labels(), c("setup", "appendix-code", "citations")), echo=TRUE, eval=FALSE}
```
