---
title: "Predicting Activity Type and Running Performance from Strava Wearable-Tracker Data: A Supervised Learning Approach Using kNN, CART, and Linear Regression"
author: "Your Names"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Abstract
This study investigates the efficacy of supervised machine learning in the automated recognition of athletic activities using personal telemetry data. Utilizing a dataset of 105 records exported from Strava, the research aims to classify activities into their respective sport types (e.g., Run, Walk, Hike) based on four specific telemetry predictors: distance, moving time, elapsed time, and total elevation gain. The primary modeling approach utilizes the k-Nearest Neighbors (kNN) algorithm, with hyperparameters tuned via 10-fold cross-validation. Data preprocessing involves handling missing values and the critical step of feature scaling (centering and scaling) to ensure unit-independent distance calculations. The study evaluates the model's performance using a 70/30 train-test split and assesses accuracy through a confusion matrix. The findings provide insight into how basic GPS-derived metrics can serve as distinct digital signatures for different types of human movement.

# Chapter 1: Introduction


## 1.1 Background of the Study

The growth of GPS wearables and platforms like Strava has led to an increase in high-resolution athletic telemetry. These devices capture core metrics such as distance, duration, and elevation. In fitness software development, automatically categorizing these data streams into specific sport types—known as Human Activity Recognition (HAR)—is essential for features like auto-tagging and personalized training feedback (Cai et al., 2024; Arora et al., 2023).

Supervised machine learning offers a method to automate this classification. By using physical attributes as predictors, models can identify the distinct "digital signatures" that differentiate a "Run" from a "Walk" or "Hike." The k-Nearest Neighbors (kNN) algorithm is a proven method for HAR due to its effectiveness in pattern recognition (Shaban, 2025; Thongsri et al., 2021). However, because kNN calculates similarity based on distance, feature scaling is a mandatory preprocessing step to ensure that different units (e.g., meters vs. minutes) do not bias the model (De Carvalho et al., 2025; Pratama et al., 2024).

This study uses 105 activity records to test a predictive pipeline. The goal is to demonstrate how basic movement data can be used to accurately classify athletic activities without manual user input.

## 1.2 Statement of the Problem

The primary challenge addressed in this study is to determine if a computer can accurately distinguish between different athletic activities using only basic movement telemetry, without the need for manual user input.

Specifically, the study seeks to answer the following questions:
1. What are the descriptive characteristics (mean, range, and distribution) of the four primary predictors—Distance, Moving Time, Elapsed Time, and Total Elevation Gain—within this personal Strava dataset?
2. To what extent can the k-Nearest Neighbors (kNN) algorithm accurately classify the sport type using these four telemetry variables?
3. How does the accuracy of the model change across different values of k (neighbors), and which k value is optimal for this specific dataset?
4. Does the necessity of feature scaling (centering and scaling) significantly impact the model's ability to interpret variables with vastly different ranges, such as distance in meters versus elevation in meters?
5. Based on the Confusion Matrix results, which activity types are most frequently misclassified by the algorithm?

## 1.3 Objectives of the Study

### General Objective

To implement and evaluate a k-Nearest Neighbors (kNN) classification model to accurately predict athletic sport types based on Strava telemetry data.

### Specific Objectives

1. To describe the dataset through statistical summaries of the four telemetry predictors.
2. To preprocess the raw data by removing missing values and applying feature scaling (centering and scaling) to ensure the kNN algorithm treats all predictors with equal weight.
3. To partition the data into a training set (70%) and a testing set (30%) using stratified sampling.
4. To tune the kNN model by testing a sequence of k values (from 1 to 15) using 10-fold cross-validation to find the highest classification accuracy.
5. To evaluate the final model using a Confusion Matrix to report Accuracy and identify specific patterns of classification error between sport types.


## 1.4 Hypotheses

H0 : GPS telemetry data (distance, time, and elevation) provide no significant predictive power in distinguishing between athletic activity types using a kNN model.
Ha : GPS telemetry data provide significant predictive power, allowing the kNN model to achieve high classification accuracy across different sport types

## 1.5 Significance of the Study

For Athletes: Demonstrates how their "movement signature" is captured by data, highlighting which metrics (like elevation vs. time) define their various workout types.
For Data Science Students: Provides a clear, reproducible example of the kNN classification workflow, specifically emphasizing the importance of Cross-Validation and Feature Scaling in distance-based algorithms.
For App Developers: Validates the use of lightweight, supervised models for the automation of activity recognition features in mobile fitness applications.


## 1.6 Scope and Limitations

**Scope.** The study is restricted to a personal dataset of 105 activity records from a single Strava account. The analysis is strictly a Classification task using the target variable sport type. The predictors are limited to the following four telemetry metrics: distance, moving time, elapsed time, and total elevation gain.

**Limitations.**

1. With only 105 records, the model's ability to learn rare activity types (e.g., "Hike" or "Walk") may be limited compared to the dominant "Run" category.
2. The findings are specific to the movement patterns of one individual and may not generalize to the broader population.
3. The study does not include heart rate, cadence, or weather data, which are often used in more complex industry-level classification models.


# Chapter 2: Review of Related Literature

> This chapter is to be written by the researchers. The five suggested article clusters below should guide the search.

## 2.1 Suggested Article Clusters

1. **Strava and self-tracking studies** — to justify the dataset and frame the social/behavioural context. *Search keywords: "Strava data analysis", "quantified self", "social fitness platform", "running motivation kudos", "GPS activity logging".*
2. **Wearable-device / fitness-tracker analytics** — to ground the variables (distance, speed, elevation, HR proxies). *Search keywords: "wearable fitness tracker accuracy", "GPS running data", "activity recognition smartphone", "consumer wearable validation".*
3. **Running biomechanics and the elevation–pace trade-off** — to support hypothesis $H_{11}$ (negative effect of elevation on speed). *Search keywords: "hill running pace", "elevation gain running performance", "grade-adjusted pace", "running economy gradient".*
4. **Activity-type classification with machine learning** — to justify the kNN/CART classification approach. *Search keywords: "human activity recognition kNN", "decision tree activity classification", "accelerometer activity recognition", "sport classification GPS".*
5. **Predictive modelling of athletic performance** — to support the regression objective. *Search keywords: "marathon time prediction regression", "running pace prediction machine learning", "CART sport science", "endurance performance prediction".*

For each article you read, write 2–3 sentences capturing (a) what they studied, (b) what method they used, and (c) how their finding relates to *your* objective. Aim for 8–12 articles total.

## 2.2 Related Studies

*Summarize previous empirical research here.*

## 2.3 Related Literature

*Discuss the underlying theories and concepts (e.g., supervised learning theory, distance metrics for kNN, recursive partitioning for CART, ordinary-least-squares assumptions).*

## 2.4 Conceptual Framework

```
[ Independent Variables ]                       [ Dependent Variables ]
distance, moving_time,           +--- kNN / CART --------> Activity Type (categorical)
elapsed_time, elevation_gain,    |
max_speed, athlete_count   ------+
                                 +--- Lin. Reg. / Reg.Tree --> Average Speed (numeric)
```

```{r conceptual-framework, echo=FALSE}
# A simple visual placeholder; replace with a proper diagram if desired.
plot.new()
title("Conceptual Framework: Predictors -> Models -> Outcomes")
text(0.5, 0.7,
     "distance, moving_time, elapsed_time,\nelevation_gain, max_speed, athlete_count",
     cex = 0.9)
arrows(0.5, 0.6, 0.3, 0.4); arrows(0.5, 0.6, 0.7, 0.4)
text(0.3, 0.35, "kNN / CART", cex = 0.9, font = 2)
text(0.7, 0.35, "Lin. Reg. / Reg.Tree", cex = 0.9, font = 2)
arrows(0.3, 0.30, 0.3, 0.15); arrows(0.7, 0.30, 0.7, 0.15)
text(0.3, 0.10, "Activity Type", cex = 0.9)
text(0.7, 0.10, "Average Speed", cex = 0.9)
```

# Chapter 3: Methodology

## 3.1 Research Design

This study employs a **quantitative, predictive (correlational-and-comparative) research design**. It is *quantitative* because all variables are numerical or categorical with finite levels; it is *predictive* because the primary aim is to forecast unseen outcomes (activity type, average speed) rather than merely describe; and it is *comparative* because two algorithms are pitted against each other for each task (kNN vs. CART for classification, multiple linear regression vs. regression tree for speed prediction).

## 3.2 Data Collection

The dataset was obtained from a public **Strava** athlete export, distributed via Kaggle as `Strava Running Data.csv`. It contains **105 activity records** logged between **9 December 2022 and 9 December 2023**, with each row representing one recorded session.

```{r load-data}
library(tidyverse)
# use here() to locate the dataset reliably when knitting from a different wd
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
library(here)
strava <- read.csv(here::here("Strava Running Data.csv"), stringsAsFactors = FALSE)
dim(strava)
glimpse(strava)
```

## 3.3 Variables Description

| Role | Variable | Type | Unit | Description |
|---|---|---|---|---|
| ID | `Sr..no.` | integer | – | Activity row identifier |
| Date | `start_date_local` | datetime | ISO-8601 | Activity start time, athlete's local TZ |
| **Target (Classification)** | `type` / `sport_type` | factor | – | Run, Walk, Hike, Workout (Tennis) |
| **Target (Regression)** | `average_speed` | numeric | m/s | Mean speed during moving time |
| Predictor | `distance` | numeric | m | Total distance covered |
| Predictor | `moving_time` | numeric | s | Time actually moving |
| Predictor | `elapsed_time` | numeric | s | Total elapsed clock time |
| Predictor | `total_elevation_gain` | numeric | m | Cumulative climb |
| Predictor | `max_speed` | numeric | m/s | Peak speed during activity |
| Engagement | `achievement_count`, `kudos_count`, `comment_count`, `athlete_count`, `photo_count` | integer | – | Social/engagement metrics |
| Geo (descriptive only) | `start_latlng`, `end_latlng`, `timezone` | text | – | Location and TZ |

**Derived variables** that will be created during preprocessing:

- `pace_min_per_km = (moving_time/60) / (distance/1000)` — easier to interpret than m/s for runners.
- `is_run = factor(type == "Run")` — for an optional binary classification version, used to address class imbalance.
- `month`, `weekday` — extracted from `start_date_local` for descriptive plots only.

## 3.4 Data Preprocessing

Before modelling, a data-fitness diagnostic is run (see Appendix A: `data_fitness_check.R`) to confirm that the dataset is suitable for kNN and CART. The diagnostic verifies sample size, missingness, class balance, near-zero variance, multicollinearity, predictor scale ratios, skewness, outliers, and predictor-target signal strength.

```{r preprocess}
library(caret)

# 1. Drop rows with distance == 0 (e.g., the Tennis "Workout" row)
strava <- strava %>% filter(distance > 0)

# 2. Drop the obviously-erroneous elapsed_time outlier
#    (62 993 s for a 1 304 m run is logging noise, not signal)
strava <- strava %>% filter(elapsed_time < 30000)

# 3. Convert types
strava$type       <- factor(strava$type)
strava$sport_type <- factor(strava$sport_type)

# 4. Keep modelling variables only
model_df <- strava %>%
  select(type, distance, moving_time, elapsed_time,
         total_elevation_gain, average_speed, max_speed,
         athlete_count)

dim(model_df)
```

## 3.5 Train–Test Split

```{r split}
set.seed(123)
idx       <- createDataPartition(model_df$type, p = 0.7, list = FALSE)
train_set <- model_df[idx, ]
test_set  <- model_df[-idx, ]

cat("Train rows:", nrow(train_set), " Test rows:", nrow(test_set), "\n")
```

## 3.6 Data Analysis Techniques

### 3.6.1 Descriptive Statistics

Functions `summary()` and `psych::describe()` will report mean, SD, median, IQR, skewness, and kurtosis for every numeric variable. Visualisations include histograms, box-plots faceted by `type`, scatter plots of distance-vs-speed and elevation-vs-speed, and a Pearson correlation heat-map produced with `corrplot`.

### 3.6.2 Inferential Statistics

- **Pearson correlation** between `total_elevation_gain` and `average_speed` (runs only) tests $H_{01}$.
- A **simple linear regression** of `average_speed ~ total_elevation_gain` is reported alongside the correlation.
- **McNemar's test** is applied to the test-set predictions of kNN vs. CART for $H_{02}$.
- A **paired t-test** on absolute residuals of the linear regression vs. the regression tree tests $H_{03}$.

### 3.6.3 Supervised Machine-Learning Models

**A. kNN Classifier (Activity Type)**

```{r knn}
library(class)
num_cols <- c("distance","moving_time","elapsed_time",
              "total_elevation_gain","average_speed","max_speed",
              "athlete_count")
preProc  <- preProcess(train_set[, num_cols], method = c("center","scale"))
train_s  <- predict(preProc, train_set[, num_cols])
test_s   <- predict(preProc, test_set[, num_cols])

# Tune k via 10-fold cross-validation
ctrl <- trainControl(method = "cv", number = 10)
knn_tune <- train(x = train_s, y = train_set$type,
                  method = "knn", trControl = ctrl,
                  tuneGrid = expand.grid(k = seq(1, 15, 2)))
knn_pred <- predict(knn_tune, test_s)
confusionMatrix(knn_pred, test_set$type)
```

**B. CART Classifier (Activity Type)**

```{r cart-class}
library(rpart); library(rpart.plot)
cart_cls <- rpart(type ~ distance + moving_time + elapsed_time +
                  total_elevation_gain + average_speed + max_speed +
                  athlete_count,
                  data = train_set, method = "class")
rpart.plot(cart_cls, type = 2, extra = 104,
           main = "CART: Activity-Type Classification Tree")
cart_pred <- predict(cart_cls, test_set, type = "class")
confusionMatrix(cart_pred, test_set$type)
```

**C. Multiple Linear Regression (Average Speed, Runs only)**

```{r lm}
runs <- model_df %>% filter(type == "Run")
set.seed(123)
ridx    <- createDataPartition(runs$average_speed, p = 0.7, list = FALSE)
r_train <- runs[ridx, ]
r_test  <- runs[-ridx, ]

lm_fit <- lm(average_speed ~ distance + moving_time +
             total_elevation_gain + athlete_count, data = r_train)
summary(lm_fit)
lm_pred <- predict(lm_fit, r_test)
postResample(lm_pred, r_test$average_speed)   # RMSE, R^2, MAE
```

**D. Regression Tree (Average Speed, Runs only)**

```{r rtree}
rt_fit  <- rpart(average_speed ~ distance + moving_time +
                 total_elevation_gain + athlete_count,
                 data = r_train, method = "anova")
rpart.plot(rt_fit, type = 2,
           main = "Regression Tree: Predicting Average Running Speed")
rt_pred <- predict(rt_fit, r_test)
postResample(rt_pred, r_test$average_speed)
```

### 3.6.4 Model-Evaluation Metrics

| Task | Metric | Tool / Function |
|---|---|---|
| Classification | Accuracy, Sensitivity, Specificity, Kappa | `caret::confusionMatrix()` |
| Classification | Class-imbalance-robust comparison | McNemar's test |
| Regression | R-squared, RMSE, MAE | `caret::postResample()` |
| Regression | Residual normality / homoscedasticity | `plot(lm_fit)` diagnostics |

### 3.6.5 Software and Reproducibility

All analysis is performed in **R 4.x** within **RStudio** using **R Markdown**. Required packages: `tidyverse`, `caret`, `class`, `rpart`, `rpart.plot`, `corrplot`, `psych`, `e1071`, `ggplot2`. The random seed `set.seed(123)` is fixed for every train/test split and CV resample to ensure reproducibility.

# Chapter 4: Results and Discussion

> *Placeholder — to be completed after running the analysis.*

## 4.1 Descriptive Statistics

```{r}
# library(psych); describe(model_df)
# summary(model_df)
```

## 4.2 Visualisation

```{r}
library(ggplot2)
# ggplot(model_df, aes(x = total_elevation_gain, y = average_speed, color = type)) +
#   geom_point() + theme_minimal() +
#   labs(title = "Elevation Gain vs. Average Speed by Activity Type")
```

## 4.3 Model Results and Interpretation

*Discuss confusion-matrix results, k tuning, tree splits, regression coefficients, and which model wins each task.*

# Chapter 5: Conclusion and Recommendations

## 5.1 Conclusion

*Summarize key findings against each Specific Objective and each hypothesis decision.*

## 5.2 Recommendations

*Practical suggestions for athletes, app developers, and future researchers.*

# References

*Use APA 7 format. Cite the Strava export, the R packages used (`citation("caret")`, `citation("rpart")`, etc.), and the 8–12 articles from your literature review.*

# Appendices

## Appendix A: Data-Fitness Diagnostic Script

The script `data_fitness_check.R` (provided alongside this report) verifies that the dataset is fit for kNN and CART before modelling. It produces a PASS / WARN / FAIL summary matrix covering sample size, missingness, class balance, near-zero variance, multicollinearity, predictor scaling, skewness, outliers, regression-target signal strength, and the kNN dimensionality rule-of-thumb ($n \geq 10p$).

To run it:

```{r data-fitness, eval=FALSE}
source("data_fitness_check.R")
```

## Appendix B: Full Code

*(All chunks above are included by `echo = TRUE` in the global setup.)*
