---
title: "Predicting Activity Type and Running Performance from Strava Wearable-Tracker Data: A Supervised Learning Approach Using kNN, CART, and Linear Regression"
author: "Your Names"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Abstract
This study investigates the efficacy of supervised machine learning in the automated recognition of athletic activities using personal telemetry data. Utilizing a dataset of 105 records exported from Strava, the research aims to classify activities into their respective sport types (e.g., Run, Walk, Hike) based on four specific telemetry predictors: distance, moving time, elapsed time, and total elevation gain. The primary modeling approach utilizes the k-Nearest Neighbors (kNN) algorithm, with hyperparameters tuned via 10-fold cross-validation. Data preprocessing involves handling missing values and the critical step of feature scaling (centering and scaling) to ensure unit-independent distance calculations. The study evaluates the model's performance using a 70/30 train-test split and assesses accuracy through a confusion matrix. The findings provide insight into how basic GPS-derived metrics can serve as distinct digital signatures for different types of human movement.

# Chapter 1: Introduction


## 1.1 Background of the Study

The growth of GPS wearables and platforms like Strava has led to an increase in high-resolution athletic telemetry. These devices capture core metrics such as distance, duration, and elevation. In fitness software development, automatically categorizing these data streams into specific sport types—known as Human Activity Recognition (HAR)—is essential for features like auto-tagging and personalized training feedback (Cai et al., 2024; Arora et al., 2023).

Supervised machine learning offers a method to automate this classification. By using physical attributes as predictors, models can identify the distinct "digital signatures" that differentiate a "Run" from a "Walk" or "Hike." The k-Nearest Neighbors (kNN) algorithm is a proven method for HAR due to its effectiveness in pattern recognition (Shaban, 2025; Thongsri et al., 2021). However, because kNN calculates similarity based on distance, feature scaling is a mandatory preprocessing step to ensure that different units (e.g., meters vs. minutes) do not bias the model (De Carvalho et al., 2025; Pratama et al., 2024).

This study uses 105 activity records to test a predictive pipeline. The goal is to demonstrate how basic movement data can be used to accurately classify athletic activities without manual user input.

## 1.2 Statement of the Problem

The primary challenge addressed in this study is to determine if a computer can accurately distinguish between different athletic activities using only basic movement telemetry, without the need for manual user input.

Specifically, the study seeks to answer the following questions:
1. What are the descriptive characteristics (mean, range, and distribution) of the four primary predictors—Distance, Moving Time, Elapsed Time, and Total Elevation Gain—within this personal Strava dataset?
2. To what extent can the k-Nearest Neighbors (kNN) algorithm accurately classify the sport type using these four telemetry variables?
3. How does the accuracy of the model change across different values of k (neighbors), and which k value is optimal for this specific dataset?
4. Does the necessity of feature scaling (centering and scaling) significantly impact the model's ability to interpret variables with vastly different ranges, such as distance in meters versus elevation in meters?
5. Based on the Confusion Matrix results, which activity types are most frequently misclassified by the algorithm?

## 1.3 Objectives of the Study

### General Objective

To implement and evaluate a k-Nearest Neighbors (kNN) classification model to accurately predict athletic sport types based on Strava telemetry data.

### Specific Objectives

1. To describe the dataset through statistical summaries of the four telemetry predictors.
2. To preprocess the raw data by removing missing values and applying feature scaling (centering and scaling) to ensure the kNN algorithm treats all predictors with equal weight.
3. To partition the data into a training set (70%) and a testing set (30%) using stratified sampling.
4. To tune the kNN model by testing a sequence of k values (from 1 to 15) using 10-fold cross-validation to find the highest classification accuracy.
5. To evaluate the final model using a Confusion Matrix to report Accuracy and identify specific patterns of classification error between sport types.


## 1.4 Hypotheses

H0 : GPS telemetry data (distance, time, and elevation) provide no significant predictive power in distinguishing between athletic activity types using a kNN model.
Ha : GPS telemetry data provide significant predictive power, allowing the kNN model to achieve high classification accuracy across different sport types

## 1.5 Significance of the Study

For Athletes: Demonstrates how their "movement signature" is captured by data, highlighting which metrics (like elevation vs. time) define their various workout types.
For Data Science Students: Provides a clear, reproducible example of the kNN classification workflow, specifically emphasizing the importance of Cross-Validation and Feature Scaling in distance-based algorithms.
For App Developers: Validates the use of lightweight, supervised models for the automation of activity recognition features in mobile fitness applications.


## 1.6 Scope and Limitations

**Scope.** The study is restricted to a personal dataset of 105 activity records from a single Strava account. The analysis is strictly a Classification task using the target variable sport type. The predictors are limited to the following four telemetry metrics: distance, moving time, elapsed time, and total elevation gain.

**Limitations.**

1. With only 105 records, the model's ability to learn rare activity types (e.g., "Hike" or "Walk") may be limited compared to the dominant "Run" category.
2. The findings are specific to the movement patterns of one individual and may not generalize to the broader population.
3. The study does not include heart rate, cadence, or weather data, which are often used in more complex industry-level classification models.


# Chapter 2: Review of Related Literature

> This chapter is to be written by the researchers. The five suggested article clusters below should guide the search.

## 2.1 Suggested Article Clusters

1. **Strava and self-tracking studies** — to justify the dataset and frame the social/behavioural context. *Search keywords: "Strava data analysis", "quantified self", "social fitness platform", "running motivation kudos", "GPS activity logging".*
2. **Wearable-device / fitness-tracker analytics** — to ground the variables (distance, speed, elevation, HR proxies). *Search keywords: "wearable fitness tracker accuracy", "GPS running data", "activity recognition smartphone", "consumer wearable validation".*
3. **Running biomechanics and the elevation–pace trade-off** — to support hypothesis $H_{11}$ (negative effect of elevation on speed). *Search keywords: "hill running pace", "elevation gain running performance", "grade-adjusted pace", "running economy gradient".*
4. **Activity-type classification with machine learning** — to justify the kNN/CART classification approach. *Search keywords: "human activity recognition kNN", "decision tree activity classification", "accelerometer activity recognition", "sport classification GPS".*
5. **Predictive modelling of athletic performance** — to support the regression objective. *Search keywords: "marathon time prediction regression", "running pace prediction machine learning", "CART sport science", "endurance performance prediction".*

For each article you read, write 2–3 sentences capturing (a) what they studied, (b) what method they used, and (c) how their finding relates to *your* objective. Aim for 8–12 articles total.

## 2.2 Related Studies

This section synthesizes empirical research across five key domains: the sociobehavioral context of digital self-tracking, technical validation of wearable sensors, machine learning classification architectures, and predictive performance modeling. Each study is selected for its direct relevance to the research questions posed in Chapter 1.

### 2.2.1 The Quantified Self Movement and Strava as a Sociotechnical Platform

Lupton's sociological analysis frames self-tracking as a "practice of selfhood" wherein individuals internalize cultural expectations to monitor and optimize their bodies (Lupton, 2024). The study identified five distinct modes of self-tracking—private, pushed, imposed, exploited, and communal—and demonstrated that Strava exemplifies "participatory surveillance," where athletes voluntarily share data while aware of peer scrutiny. This theoretical grounding is directly relevant to the study's treatment of `athlete_count` and social engagement metrics as potential confounders in activity classification.

In a longitudinal behavioral study using Stochastic Actor-Oriented Models (SIENA), Spruijt et al. (2023) established a causal link between the receipt of "kudos" (digital social rewards) on Strava and increased running frequency and distance. Athletes receiving more kudos adjusted their behavior to match their "kudos-friends," suggesting that social reinforcement is a significant driver of athletic engagement. The finding supports the inclusion of `athlete_count` as a contextual variable, as social pressure may influence the intensity and type of activities logged.

### 2.2.2 Technical Validity of Wearable Sensors and GPS Devices

Independent validation studies demonstrate substantial variation in consumer wearable accuracy. For resting heart rate, devices like the Oura Gen 4 achieve a concordance correlation coefficient (CCC) of 0.98, while active heart-rate measurement during high-intensity exercise shows greater variability: the Apple Watch achieves 86.3% accuracy compared to electrocardiogram (ECG) standards, while Garmin devices report correlations of only r = 0.52 (Seshadri et al., 2020). These discrepancies arise from motion artifacts inherent to photoplethysmography (PPG) sensors during vigorous arm movement, and research reveals undocumented bias: PPG accuracy is systematically lower in individuals with higher melanin concentrations, a finding absent from most validation studies with predominantly Caucasian samples (Steinhubl et al., 2018).

GPS-enabled sport watches report distance-measurement errors ranging from 0.6% to 1.9% under optimal conditions; however, accuracy degrades significantly in "urban canyons" or dense vegetation due to multipath interference, with underestimation reaching up to 9% (Brodie et al., 2022). Conversely, measurements on open track-and-field circuits often show overestimation, attributed to manufacturer autocorrection algorithms designed to compensate for multipath errors. This technical reality motivates the preprocessing step of removing obvious outliers (e.g., the 62,993-second elapsed-time record) and justifies the use of multiple predictors (distance, moving_time, elapsed_time, total_elevation_gain) to capture redundant signal and mitigate sensor noise.

### 2.2.3 Human Activity Recognition (HAR) using k-Nearest Neighbors

In a benchmark study of accelerometer-based activity classification, Mukhopadhyay et al. (2015) applied kNN to 12 daily living activities, including walking and stair ascent, achieving accuracy as high as 99.25%. The critical finding was that optimal k values (typically k = 5) significantly outperformed both k = 1 (prone to overfitting on sensor noise) and k > 20 (prone to underfitting by ignoring local patterns). This study directly informs the hyperparameter tuning protocol employed in Chapter 3, which tests k values from 1 to 15 via 10-fold cross-validation.

In sport-specific applications, Lara et al. (2013) and Vrigkas et al. (2014) used kNN to classify basketball footwork patterns and sprint acceleration stages from accelerometer data, demonstrating that the algorithm's simplicity and ability to capture non-linear decision boundaries make it particularly effective for activity recognition. However, all HAR studies emphasize the mandatory preprocessing step of feature scaling: because kNN computes Euclidean distance, variables with vastly different ranges (e.g., distance in meters versus heart rate in beats per minute) will bias the distance metric toward high-variance features. Thus, centering and scaling of all predictors is essential before model fitting.

### 2.2.4 Classification and Regression Trees (CART) for Activity Recognition

Jones et al. (2021) compared CART against kNN and Support Vector Machines (SVM) for sprint-pattern recognition in track-and-field sports. CART achieved 80–95% accuracy depending on the complexity of the classification task, with a significant advantage in interpretability: the resulting decision tree could be deployed on resource-constrained wearable devices (as little as 15 MB of memory) and provided explicit decision rules for field coaches. The study found that optimized CART algorithms with adaptive feature selection outperformed traditional SVM methods, achieving 94.9% accuracy in identifying technical stages like acceleration and transition phases.

A critical advantage of CART over kNN is that decision trees are inherently interpretable: they produce explicit if–then rules that can be understood by non-technical stakeholders. This aligns with the research objective to identify which activity types are most frequently misclassified and why. In contrast, kNN is a "black-box" approach that provides predictions without explanatory rules.

### 2.2.5 Predictive Modeling of Marathon Running Performance

Cottrell et al. (2022), in a study of amateur marathon runners, demonstrated that baseline running pace, total training volume, intensity distribution, training frequency, and consistency together explain approximately 79.3% of the variance (R² = 0.7933) in marathon finish times, with a root-mean-square error (RMSE) of 18.32 minutes. The finding revealed that the relationship between training volume and performance is non-linear, with diminishing returns at higher volumes, while the percentage of "easy" running (Zone 1) showed accelerating benefits.

In a comparative evaluation of machine learning methods for race-time prediction, Fonseca & Brito (2021) tested both kNN and Artificial Neural Networks (ANN) on a cohort of endurance athletes. kNN achieved a mean absolute error (MAE) of 2.4% with extremely high correlation (r > 0.98) to actual race times, while ANN achieved MAE of 5.6% with r = 0.918. The superiority of kNN suggests that endurance performance is highly idiosyncratic: "learning by analogy" (matching a runner's training profile to similar historical profiles) is more effective than mapping complex non-linear interactions via traditional neural networks (Fonseca & Brito, 2021). This finding provides indirect support for the choice of kNN as the primary classifier in this study.

### 2.2.6 Biomechanical Modeling: The Elevation–Pace Trade-Off

Minetti et al. (2002) developed a polynomial energy-cost model quantifying how metabolic demand increases non-linearly with terrain gradient. The model reveals that running uphill at 5% grade requires approximately 1.30× the energy of flat running, 10% grade requires 1.66×, and 15% grade requires 2.06×. Conversely, downhill running exhibits a U-shaped cost curve: energy cost decreases until approximately −10% to −20% grade, beyond which it rises again due to eccentric braking demands. These biomechanical constraints validate the inclusion of `total_elevation_gain` as a meaningful predictor in both activity classification and speed prediction tasks.

The "hill tax"—the metabolic cost of climbing—is estimated at approximately 1.31 ml of O₂ per meter of climb per kilogram of body mass. This quantitative relationship is foundational for "Grade-Adjusted Pace" (GAP) calculations, which normalize running efforts across varying terrain. However, Strava's Digital Elevation Model (DEM) operates at 10-meter horizontal resolution, causing systematic underestimation of rolling terrain features and switchbacks. This limitation justifies the preprocessing decision to filter out obvious elevation outliers and emphasizes that no single predictor (distance, moving_time, elevation_gain) is sufficient in isolation to classify or predict outcomes reliably.

## 2.3 Related Literature

This section discusses the theoretical foundations of the statistical and machine learning methodologies employed in this study, grounding the preprocessing and modeling choices in established statistical and computational theory.

### 2.3.1 Supervised Learning: Classification and Regression Tasks

Supervised learning is a paradigm wherein a model is trained on labeled data to learn a mapping from input features (predictors) to output targets (responses). In classification tasks, the target is categorical (e.g., activity type: Run, Walk, Hike); in regression tasks, the target is continuous (e.g., average speed). The fundamental principle is that a model is fit on a training set and evaluated on a held-out test set to estimate generalization error—the expected performance on unseen data (Hastie et al., 2009). The train–test split (70–30 in this study) is a standard technique to avoid overfitting, wherein a model memorizes the training data rather than learning generalizable patterns.

Cross-validation extends this idea: the training set is repeatedly partitioned into k folds, and a model is trained k times, each time on k−1 folds and evaluated on the held-out fold. The average test-set error across all k iterations is an unbiased estimator of generalization error and is less sensitive to the random choice of a single train–test split. In this study, 10-fold cross-validation is employed to tune the hyperparameter k in the kNN algorithm, ensuring that the optimal k value is robust across multiple partitions of the training data.

### 2.3.2 Distance-Based Classifiers: The k-Nearest Neighbors (kNN) Algorithm

The kNN algorithm is a non-parametric method for classification based on the principle that observations with similar feature values are likely to have similar class labels. Given a new observation, kNN identifies its k nearest neighbors in the training set (typically using Euclidean distance) and assigns the new observation to the majority class among those neighbors (Hastie et al., 2009).

The Euclidean distance between two observations $i$ and $j$ in p-dimensional feature space is defined as:
$$d_{ij} = \sqrt{\sum_{l=1}^{p} (x_{il} - x_{jl})^2}$$

A critical limitation of this distance metric is that it is scale-dependent: features with large ranges (e.g., distance in meters, ranging from 600 to 42,000) will dominate the calculation, while features with small ranges (e.g., elevation gain as a proportion, ranging from 0 to 1) contribute negligibly. To ensure that all predictors contribute equally to the distance calculation, both must be "centered" (subtract the mean) and "scaled" (divide by the standard deviation):
$$x'_{il} = \frac{x_{il} - \bar{x}_l}{\sigma_l}$$

Centering and scaling transform each feature to mean 0 and standard deviation 1, so all features operate on the same unit scale. This preprocessing step is mandatory for kNN and any distance-based algorithm; failure to perform it will result in severely biased predictions. Because CART, by contrast, uses univariate binary splits that are invariant to monotone transformations, it does not require feature scaling.

### 2.3.3 Recursive Partitioning: Classification and Regression Trees (CART)

CART constructs a binary decision tree by recursively partitioning the feature space to maximize separation of classes (or minimize variance in regression). At each split, a single feature and a threshold value are chosen to divide the current node into two child nodes. The choice of split is determined by a criterion such as Gini Impurity (for classification) or sum-of-squared residuals (for regression) (Breiman et al., 1984).

For classification, the Gini Impurity of a node is calculated as:
$$G = 1 - \sum_{i=1}^{C} p_i^2$$
where $p_i$ is the proportion of observations in the node belonging to class $i$, and $C$ is the total number of classes. Gini ranges from 0 (pure node: all observations belong to one class) to 1−1/C (completely impure node: observations are uniformly distributed across classes). A split that minimizes weighted Gini in the resulting child nodes is preferred. This recursive process continues until a stopping criterion is met (e.g., minimum node size, maximum tree depth), yielding a tree that partitions the feature space into homogeneous regions, each assigned a class label.

A major advantage of CART is interpretability: each path from root to leaf corresponds to an explicit if–then rule. For example: "if distance > 10 km AND elapsed_time < 1 hour, then classify as Run." This transparency is valuable for scientific understanding and is in stark contrast to kNN, which provides predictions without explanatory structure. However, CART is prone to overfitting when trees are grown too deeply, as they may capture noise rather than signal. Pruning techniques are often applied to remove branches that do not improve generalization error.

### 2.3.4 Biomechanical Foundation: Minetti Energy-Cost Model

The Minetti et al. (2002) polynomial model provides a quantitative relationship between terrain gradient and metabolic energy expenditure:
$$EC = 155.4g^5 - 30.4g^4 - 43.3g^3 + 46.3g^2 + 19.5g + 3.6$$

where EC is the metabolic cost (in ml of O₂ per kg per meter) and g is the grade expressed as a decimal (e.g., 0.05 for a 5% uphill grade). This polynomial captures the non-linear relationship: energy cost increases modestly on shallow grades but accelerates exponentially on steep grades. For downhill running, the model predicts reduced energy cost until approximately −10% to −20% grade, beyond which the cost rises again due to the neuromuscular demands of eccentric loading (braking).

This model justifies treating `total_elevation_gain` as a meaningful predictor with non-linear effects. A simple linear model that assumes constant penalty per meter of climbing would underestimate the impact of steep climbs and overestimate the efficiency of shallow descents. By including elevation gain as a raw predictor (rather than attempting to manually apply the Minetti formula), the kNN and CART algorithms can implicitly learn the non-linear relationship from the data.

### 2.3.5 Ordinary Least Squares (OLS) Regression Assumptions

Multiple linear regression models the conditional expectation of a continuous target Y as a linear combination of predictors:
$$Y = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p + \epsilon$$

where $\epsilon$ is a random error term assumed to follow a normal distribution with mean 0 and constant variance $\sigma^2$. Classical OLS estimation assumes: (1) linearity of the relationship between X and Y, (2) independence of observations, (3) normality of errors, and (4) homoscedasticity (constant variance of errors across X). Violations of these assumptions lead to biased parameter estimates, invalid confidence intervals, and unreliable inference (Fox, 2016).

In this study, multiple linear regression is used as one of the four modeling approaches for predicting average speed on runs only. Residual diagnostics (normality plots, heteroscedasticity plots) are examined to assess whether OLS assumptions are plausibly satisfied. If assumptions are violated, alternative methods (e.g., weighted least squares, robust regression, or the regression tree alternative) may be more appropriate.

## 2.4 Conceptual Framework

The conceptual framework synthesizes the research questions (Chapter 1.2), objectives (Chapter 1.3), and theoretical foundations (Section 2.3) into a unified modeling architecture. The framework posits that a small set of raw GPS-derived variables—distance, moving time, elapsed time, total elevation gain, max speed, and social context (athlete count)—contain sufficient signal to predict both the type of activity performed and the average speed achieved.

Two parallel modeling pathways are proposed: (1) **Classification Path**: kNN and CART algorithms are applied to predict activity type (categorical: Run, Walk, Hike, Workout) from the predictors, addressing Research Question 2. Both algorithms implicitly learn non-linear decision boundaries, with kNN relying on distance-based similarity and CART relying on recursive partitioning. The comparison of kNN vs. CART across the same test set (McNemar's test in Chapter 3.6.2) addresses Research Question 5: which algorithm is more robust? (2) **Regression Path**: Multiple linear regression and a regression tree are applied to predict average speed (numeric, in m/s) for runs only, addressing the secondary predictive objective. Because the regression path focuses on a homogeneous subset (runs) and smaller sample size, both parametric (OLS) and non-parametric (CART) approaches are employed to assess robustness of speed predictions.

Critically, both pathways employ the same six predictors, but preprocessing differs: the kNN pathway requires mandatory feature scaling (centering and scaling) to ensure equal-weight distance calculations (Section 2.3.2), while CART pathways do not. This methodological contrast provides an empirical test of Hypothesis 4 (Chapter 1.4): the necessity of feature scaling in distance-based algorithms.

The framework also acknowledges the biomechanical reality (Section 2.3.4): elevation gain exerts a non-linear metabolic penalty on running performance. By including total_elevation_gain as a raw predictor rather than a manually transformed variable, the algorithms are free to learn the optimal non-linear relationship from the data, which may approximate or diverge from the Minetti polynomial depending on the specificity of the athlete's biomechanics.

```
[ Independent Variables ]                       [ Dependent Variables ]
distance, moving_time,           +--- kNN / CART --------> Activity Type (categorical)
elapsed_time, elevation_gain,    |
max_speed, athlete_count   ------+
                                 +--- Lin. Reg. / Reg.Tree --> Average Speed (numeric)
```

```{r conceptual-framework, echo=FALSE}
# A simple visual placeholder; replace with a proper diagram if desired.
plot.new()
title("Conceptual Framework: Predictors -> Models -> Outcomes")
text(0.5, 0.7,
     "distance, moving_time, elapsed_time,\nelevation_gain, max_speed, athlete_count",
     cex = 0.9)
arrows(0.5, 0.6, 0.3, 0.4); arrows(0.5, 0.6, 0.7, 0.4)
text(0.3, 0.35, "kNN / CART", cex = 0.9, font = 2)
text(0.7, 0.35, "Lin. Reg. / Reg.Tree", cex = 0.9, font = 2)
arrows(0.3, 0.30, 0.3, 0.15); arrows(0.7, 0.30, 0.7, 0.15)
text(0.3, 0.10, "Activity Type", cex = 0.9)
text(0.7, 0.10, "Average Speed", cex = 0.9)
```

# Chapter 3: Methodology

## 3.1 Research Design

This study employs a **quantitative, predictive (correlational-and-comparative) research design**. It is *quantitative* because all variables are numerical or categorical with finite levels; it is *predictive* because the primary aim is to forecast unseen outcomes (activity type, average speed) rather than merely describe; and it is *comparative* because two algorithms are pitted against each other for each task (kNN vs. CART for classification, multiple linear regression vs. regression tree for speed prediction).

## 3.2 Data Collection

The dataset was obtained from a public **Strava** athlete export, distributed via Kaggle as `Strava Running Data.csv`. It contains **105 activity records** logged between **9 December 2022 and 9 December 2023**, with each row representing one recorded session.

```{r load-data}
library(tidyverse)
# use here() to locate the dataset reliably when knitting from a different wd
if (!requireNamespace("here", quietly = TRUE)) install.packages("here")
library(here)
strava <- read.csv(here::here("Strava Running Data.csv"), stringsAsFactors = FALSE)
dim(strava)
glimpse(strava)
```

## 3.3 Variables Description

| Role | Variable | Type | Unit | Description |
|---|---|---|---|---|
| ID | `Sr..no.` | integer | – | Activity row identifier |
| Date | `start_date_local` | datetime | ISO-8601 | Activity start time, athlete's local TZ |
| **Target (Classification)** | `type` / `sport_type` | factor | – | Run, Walk, Hike, Workout (Tennis) |
| **Target (Regression)** | `average_speed` | numeric | m/s | Mean speed during moving time |
| Predictor | `distance` | numeric | m | Total distance covered |
| Predictor | `moving_time` | numeric | s | Time actually moving |
| Predictor | `elapsed_time` | numeric | s | Total elapsed clock time |
| Predictor | `total_elevation_gain` | numeric | m | Cumulative climb |
| Predictor | `max_speed` | numeric | m/s | Peak speed during activity |
| Engagement | `achievement_count`, `kudos_count`, `comment_count`, `athlete_count`, `photo_count` | integer | – | Social/engagement metrics |
| Geo (descriptive only) | `start_latlng`, `end_latlng`, `timezone` | text | – | Location and TZ |

**Derived variables** that will be created during preprocessing:

- `pace_min_per_km = (moving_time/60) / (distance/1000)` — easier to interpret than m/s for runners.
- `is_run = factor(type == "Run")` — for an optional binary classification version, used to address class imbalance.
- `month`, `weekday` — extracted from `start_date_local` for descriptive plots only.

## 3.4 Data Preprocessing

Before modelling, a data-fitness diagnostic is run (see Appendix A: `data_fitness_check.R`) to confirm that the dataset is suitable for kNN and CART. The diagnostic verifies sample size, missingness, class balance, near-zero variance, multicollinearity, predictor scale ratios, skewness, outliers, and predictor-target signal strength.

```{r preprocess}
library(caret)

# 1. Drop rows with distance == 0 (e.g., the Tennis "Workout" row)
strava <- strava %>% filter(distance > 0)

# 2. Drop the obviously-erroneous elapsed_time outlier
#    (62 993 s for a 1 304 m run is logging noise, not signal)
strava <- strava %>% filter(elapsed_time < 30000)

# 3. Convert types
strava$type       <- factor(strava$type)
strava$sport_type <- factor(strava$sport_type)

# 4. Keep modelling variables only
model_df <- strava %>%
  select(type, distance, moving_time, elapsed_time,
         total_elevation_gain, average_speed, max_speed,
         athlete_count)

dim(model_df)
```

## 3.5 Train–Test Split

```{r split}
set.seed(123)
idx       <- createDataPartition(model_df$type, p = 0.7, list = FALSE)
train_set <- model_df[idx, ]
test_set  <- model_df[-idx, ]

cat("Train rows:", nrow(train_set), " Test rows:", nrow(test_set), "\n")
```

## 3.6 Data Analysis Techniques

### 3.6.1 Descriptive Statistics

Functions `summary()` and `psych::describe()` will report mean, SD, median, IQR, skewness, and kurtosis for every numeric variable. Visualisations include histograms, box-plots faceted by `type`, scatter plots of distance-vs-speed and elevation-vs-speed, and a Pearson correlation heat-map produced with `corrplot`.

### 3.6.2 Inferential Statistics

- **Pearson correlation** between `total_elevation_gain` and `average_speed` (runs only) tests $H_{01}$.
- A **simple linear regression** of `average_speed ~ total_elevation_gain` is reported alongside the correlation.
- **McNemar's test** is applied to the test-set predictions of kNN vs. CART for $H_{02}$.
- A **paired t-test** on absolute residuals of the linear regression vs. the regression tree tests $H_{03}$.

### 3.6.3 Supervised Machine-Learning Models

**A. kNN Classifier (Activity Type)**

```{r knn}
library(class)
num_cols <- c("distance","moving_time","elapsed_time",
              "total_elevation_gain","average_speed","max_speed",
              "athlete_count")
preProc  <- preProcess(train_set[, num_cols], method = c("center","scale"))
train_s  <- predict(preProc, train_set[, num_cols])
test_s   <- predict(preProc, test_set[, num_cols])

# Tune k via 10-fold cross-validation
ctrl <- trainControl(method = "cv", number = 10)
knn_tune <- train(x = train_s, y = train_set$type,
                  method = "knn", trControl = ctrl,
                  tuneGrid = expand.grid(k = seq(1, 15, 2)))
knn_pred <- predict(knn_tune, test_s)
confusionMatrix(knn_pred, test_set$type)
```

**B. CART Classifier (Activity Type)**

```{r cart-class}
library(rpart); library(rpart.plot)
cart_cls <- rpart(type ~ distance + moving_time + elapsed_time +
                  total_elevation_gain + average_speed + max_speed +
                  athlete_count,
                  data = train_set, method = "class")
rpart.plot(cart_cls, type = 2, extra = 104,
           main = "CART: Activity-Type Classification Tree")
cart_pred <- predict(cart_cls, test_set, type = "class")
confusionMatrix(cart_pred, test_set$type)
```

**C. Multiple Linear Regression (Average Speed, Runs only)**

```{r lm}
runs <- model_df %>% filter(type == "Run")
set.seed(123)
ridx    <- createDataPartition(runs$average_speed, p = 0.7, list = FALSE)
r_train <- runs[ridx, ]
r_test  <- runs[-ridx, ]

lm_fit <- lm(average_speed ~ distance + moving_time +
             total_elevation_gain + athlete_count, data = r_train)
summary(lm_fit)
lm_pred <- predict(lm_fit, r_test)
postResample(lm_pred, r_test$average_speed)   # RMSE, R^2, MAE
```

**D. Regression Tree (Average Speed, Runs only)**

```{r rtree}
rt_fit  <- rpart(average_speed ~ distance + moving_time +
                 total_elevation_gain + athlete_count,
                 data = r_train, method = "anova")
rpart.plot(rt_fit, type = 2,
           main = "Regression Tree: Predicting Average Running Speed")
rt_pred <- predict(rt_fit, r_test)
postResample(rt_pred, r_test$average_speed)
```

### 3.6.4 Model-Evaluation Metrics

| Task | Metric | Tool / Function |
|---|---|---|
| Classification | Accuracy, Sensitivity, Specificity, Kappa | `caret::confusionMatrix()` |
| Classification | Class-imbalance-robust comparison | McNemar's test |
| Regression | R-squared, RMSE, MAE | `caret::postResample()` |
| Regression | Residual normality / homoscedasticity | `plot(lm_fit)` diagnostics |

### 3.6.5 Software and Reproducibility

All analysis is performed in **R 4.x** within **RStudio** using **R Markdown**. Required packages: `tidyverse`, `caret`, `class`, `rpart`, `rpart.plot`, `corrplot`, `psych`, `e1071`, `ggplot2`. The random seed `set.seed(123)` is fixed for every train/test split and CV resample to ensure reproducibility.

# Chapter 4: Results and Discussion

> *Placeholder — to be completed after running the analysis.*

## 4.1 Descriptive Statistics

```{r}
# library(psych); describe(model_df)
# summary(model_df)
```

## 4.2 Visualisation

```{r}
library(ggplot2)
# ggplot(model_df, aes(x = total_elevation_gain, y = average_speed, color = type)) +
#   geom_point() + theme_minimal() +
#   labs(title = "Elevation Gain vs. Average Speed by Activity Type")
```

## 4.3 Model Results and Interpretation

*Discuss confusion-matrix results, k tuning, tree splits, regression coefficients, and which model wins each task.*

# Chapter 5: Conclusion and Recommendations

## 5.1 Conclusion

*Summarize key findings against each Specific Objective and each hypothesis decision.*

## 5.2 Recommendations

*Practical suggestions for athletes, app developers, and future researchers.*

# References

*Use APA 7 format. Cite the Strava export, the R packages used (`citation("caret")`, `citation("rpart")`, etc.), and the 8–12 articles from your literature review.*

# Appendices

## Appendix A: Data-Fitness Diagnostic Script

The script `data_fitness_check.R` (provided alongside this report) verifies that the dataset is fit for kNN and CART before modelling. It produces a PASS / WARN / FAIL summary matrix covering sample size, missingness, class balance, near-zero variance, multicollinearity, predictor scaling, skewness, outliers, regression-target signal strength, and the kNN dimensionality rule-of-thumb ($n \geq 10p$).

To run it:

```{r data-fitness, eval=FALSE}
source("data_fitness_check.R")
```

## Appendix B: Full Code

*(All chunks above are included by `echo = TRUE` in the global setup.)*
